### 2025-04-14 paper/main.tex
add paper draft

% !TeX program = pdflatex
% Minimally-Congested Travel Time Prediction from Sparse Open Data
% Authors: Geoff Boeing and Yuquan Zhou
% Web: https://geoffboeing.com/
% Repo: https://github.com/gboeing/travel-time-prediction

\newcommand{\myname}{Geoff Boeing and Yuquan Zhou}
\newcommand{\myemail}{boeing@usc.edu}
\newcommand{\myaffiliation}{University of Southern California}
\newcommand{\paperdate}{2025}
\newcommand{\papertitle}{Minimally-Congested Travel Time Prediction from Sparse Open Data}
\newcommand{\papercitation}{Boeing, G. and Y. Zhou. \paperdate. \papertitle. Under review at \emph{Journal Name}.}
\newcommand{\paperkeywords}{Urban Planning, Transportation, Data Science}

\RequirePackage[l2tabu,orthodox]{nag} % warn if using any obsolete or outdated commands
\documentclass[12pt,letterpaper]{article} % document style

% import encoding and font packages for pdflatex, in order
\usepackage[T1]{fontenc} % output T1 font encoding (8-bit) so accented characters are a single glyph
\usepackage[utf8]{inputenc} % allow input of utf-8 encoded characters
\usepackage{ebgaramond} % document's serif font
\usepackage{tgheros} % document's sans serif font

% import language, regionalization, microtype, and setspace in order
\usepackage[strict,autostyle]{csquotes} % smart and nestable quote marks
\usepackage[USenglish]{babel} % automatically regionalize hyphens, quote marks, etc
\usepackage{microtype} % improves text appearance with kerning, etc
\usepackage{setspace} % configure spacing between lines
\usepackage{mathtools} % use package math tool
% import everything else
\usepackage{abstract} % allow full-page title/abstract in twocolumn mode
\usepackage{amsmath}
\usepackage{authblk} % footnote-style author/affiliation info
\usepackage{booktabs} % better looking tables
\usepackage{caption} % custom figure/table caption styles
\usepackage{datetime} % enable formatting of date output
\usepackage[final]{draftwatermark} % watermark paper as a draft
\usepackage{endnotes} % enable endnotes
\usepackage{geometry} % configure page dimensions and margins
\usepackage{graphicx} % better inclusion of graphics
\usepackage{hyperref} % hypertext in document
\usepackage{natbib} % author-year citations w/ bibtex, including textual and parenthetical
\usepackage{rotating} % rotate wide tables or figures on a page to make them landscape
\usepackage{titlesec} % custom section and subsection heading
\usepackage{url} % make nice line-breakble urls
\usepackage{makecell}

% print only the month and year when using \today
\newdateformat{monthyeardate}{\monthname[\THEMONTH] \THEYEAR}

% location of figure files, via graphicx package
\graphicspath{{./figures/}}

% configure the page layout, via geometry package
\geometry{
    paper=letterpaper, % paper size
    top=3.8cm, % margin sizes
    bottom=3.8cm,
    left=4cm,
    right=4cm}
\setstretch{1} % line spacing
\clubpenalty=10000 % prevent orphans
\widowpenalty=10000 % prevent widows

% set section/subsection headings as the sans serif font, via titlesec package
\titleformat{\section}{\normalfont\sffamily\large\bfseries\color{black}}{\thesection.}{0.3em}{}
\titleformat{\subsection}{\normalfont\sffamily\small\bfseries\color{black}}{\thesubsection.}{0.3em}{}
\titleformat{\subsubsection}{\normalfont\sffamily\small\color{black}}{\thesubsubsection.}{0.3em}{}

% make figure/table captions sans-serif small font
\captionsetup{font={footnotesize,sf},labelfont=bf,labelsep=period}

% configure pdf metadata and link handling, via hyperref package
\hypersetup{
    pdfauthor={\myname},
    pdftitle={\papertitle},
    pdfsubject={\papertitle},
    pdfkeywords={\paperkeywords},
    pdffitwindow=true, % window fit to page when opened
    breaklinks=true, % break links that overflow horizontally
    colorlinks=false, % remove link color
    pdfborder={0 0 0} % remove link border
}

\begin{document}

    \title{\papertitle}
    \author[]{Redacted for review}%{\myname}
    \affil[]{Redacted for review}%{\myaffiliation}
    \date{ }%{\paperdate}

    \maketitle

    \begin{abstract}

        Travel time prediction is central to accessibility analyses, sustainable transportation infrastructure provision, and active transportation interventions. However, calculating accurate travel times, especially for driving, requires either extensive technical capacities and bespoke data, or sources like the Google Maps API that quickly becomes prohibitively expensive to analyze thousands or millions of trips necessary for metropolitan-scale planning. These obstacles make it difficult for researchers, practitioners, and community advocates to develop an evidence base for more sustainable transportation and equitable accessibility. This paper develops a free, open-source minimally-congested driving time prediction model with minimal cost, data, and computational requirements. It trains and tests this model using the Los Angeles, California urban area as a case study by calculating naïve travel times then developing a predictive random forest regression model of Google Maps' \enquote{gold standard} travel times using those naïve times plus turns and traffic controls along the naïve routes. Validation reveals that this method yields accurate, free, and simple travel time predictions.

    \end{abstract}

    \section{Introduction}

    Travel time prediction underpins many transport planning processes and research problems. It is key to understanding urban accessibility, transport mode and route choice, and location decisions. Predicting congested travel times (i.e., when traffic congestion impedes flows) requires large volumes of real-time data on traffic conditions. Such data are proprietary, expensive, and beyond the reach of many planning practitioners and scholars. This is usually not a problem for well-resourced organizations with highly-skilled labor. However, technical expertise, complicated data and software requirements, and high costs pose insurmountable obstacles for many urban planning researchers, practitioners, and community advocates.

    Although typical \enquote{naïve} travel time prediction methods that minimize travel distance or street segment traversal time provide easy solutions, they ignore delays from turns and traffic controls and thus usually under-predicts real-world travel time. However, minimally-congested travel time prediction offers a middle ground of more-accurate predictions than naïve travel times do---yet much lower data and computing requirements than traffic-aware and congested travel time prediction. Focusing on driving, this study contributes a simple, accurate, and free method of minimally-congested travel time prediction. Our results show a substantial improvement over traditional, naïve methods of travel time prediction common in the urban planning literature. As we shall show in this paper, across the Los Angeles urban area, our average trip's travel time prediction differs on average from Google's \enquote{gold standard} prediction by only 0.38 seconds. In contrast, simple---but common---naïve travel time prediction minimizing max-speed edge traversal times under-predicts Google's travel times by 182.71 seconds, on average.

    The rest of this paper is organized as follows. First it reviews the state of the art in travel time prediction and then considers the less-sophisticated methods that are standard in the current planning literature. Next it describes our data sources, models, and validation techniques. Then it presents our findings, highlighting both the feasibility and quality of our predictions in relation to simple naïve predictions or difficult expensive predictions. Finally, it concludes with a discussion of implications for research and practice.

    \section{Background}

    Modern travel time prediction techniques tend to fall into one of two camps: 1) sophisticated, accurate, but difficult to execute; or 2) naïve, less-accurate, but simple to execute. The former is the state-of-the-art and has been the subject of much work in the recent computer science and engineering literatures. The latter tends to appear often in urban planning research and practice. Here we summarize the recent advances and current standards in these fields.

    \subsection{State-of-the-Art Travel Time Prediction}

    Recent studies in the computer science and engineering literatures propose a variety of prediction techniques with heavy computing and data requirements. They usually measure their prediction accuracy by mean absolute percentage error (MAPE), defined in Equation~\ref{eq:mape}:

    \begin{equation}
        \label{eq:mape}
        \text{MAPE} = \frac{100}{n} \sum^{n}_{i=1} \left|\frac{t_i - \hat{t_i}}{t_i}\right|
    \end{equation}

    where $n$ is the total number of trips, $t_i$ is trip $i$'s observed travel time, and $\hat{t_i}$ is trip $i$'s predicted travel time. Lower MAPEs indicate higher accuracy.  These studies usually yield MAPEs within roughly 10\% of observed driving times. Some researchers use alternative measures of accuracy, such as the percentage of over- or under-prediction \citep{jenelius2013travel}, the absolute mismeasurement in minutes \citep{chiabaut2021traffic}, or degrees of predictability \citep{li2019travel}.

    These advanced travel time prediction models require massive data input as they are usually trained on millions or billions of disaggregate empirical travel time records. For example, \citet{hou2018network} use 1.5 billion GPS-based travel time records in deep learning models to predict travel time in 13 road segments in St. Louis, yielding a MAPE of 6.8\%. Using more than 100,000 trips from Windows phone GPS data, \citet{woodard2017predicting}'s technique achieves a MAPE of 10.1\% when predicting travel times. \citet{pamula2023estimation} use three months of video sensing data at a 5-minute temporal resolution in a deep learning model to predict travel times in Poland, yielding a 6.8\% MAPE. Other studies use graph neural network models to predict travel times. For example, \citet{wang2023dynamic} use data from 100,000 Chinese ride-hailing trips in graph neural network and recurrent neural network models, yielding a MAPE of 15.4\%. \citet{vankdoth2023deep} achieve 3--4\% MAPEs using Q-Traffic and TaxiBJ data from Beijing and Chengdu in deep learning and graph neural network models.

    Many of these studies focus on predicting travel times only for a specific street segment. For instance, \citet{sharmila2019svm} achieve a 10\% MAPE using GPS data in a support vector machine/particle filter model to predict travel times along specific arterials in Mumbai. \citet{chen2016multi}'s agent-based model uses data from GPS-equipped vehicles at 1-minute intervals across 123 days to achieve a MAPE under 9\% on a 95-mile freeway section in Virginia. Based on empirical travel time from GPS traces, road geometric features, and weather information, \citet{qiu2021machine} predict travel times using decision trees, random forests, extreme gradient boosting, and long short-term memory neural networks on the I-485 freeway in Charlotte, North Carolina, yielding MAPEs between 6--17\%.

    As we have seen, these studies train or validate their models with a wide variety of bespoke data sources often collected from GPS devices or proprietary sources. The Google Maps API offers another common source. Though a black-box, many researchers use Google's travel times as the \enquote{gold standard} predictions of real-world travel times, given its ubiquity and accuracy and the challenges of otherwise directly obtaining sufficient empirical travel data \citep[e.g.,][]{goudarzi2018travel,stanojevic2019mapreuse,ludwig2023traffic}.

    \subsection{Path Solving in Planning Research}

    Central to all of the preceding travel time prediction research is the concept of \textit{path solving}: identifying the shortest path between an origin and destination in a network model. In this case, the shortest path is defined as the path that minimizes travel time and could be realized methodologically in multiple ways. In the planning literature, studies variously minimize euclidean distance, network distance, network edge traversal time, free-flow travel time, or congested travel time. We discuss their trade-offs in this section.

    The actual path solving methods used in urban planning research and practice tend to differ substantially from the state of the art travel time prediction in the engineering and computer science literatures. The preceding section's models require extensive technical capacity, instrumentation, and data. In general, the planning literature's implementations of path solving tend to be simpler to execute, but potentially naïve and less-accurate. For example, many accessibility studies employ merely a (very) rough proxy for travel time by minimizing euclidean distance traveled instead \citep[e.g.,][]{macfarlane2021modeling,pearsall2020locating}. This offers the substantial benefit of simplicity (no network model, travel time date, or routing algorithms are needed) but offers a poor estimate of access in terms of travel time.

    Many other studies improve on this by instead solving shortest paths by minimizing network distance traveled. These studies often use data and tools from Esri, OpenStreetMap, or the US Census Bureau to measure network distances between origins and destinations \citep[e.g.,][]{mckenzie2020urban, jiao2021measuring, nicoletti2023disadvantaged, logan2019evaluating, tsou2005accessibility}. This offers the benefit of more-realistic distances between origins and destinations, but does not account for travel speeds. Other studies refine this by incorporating speed limit data into the network model to solve shortest paths by traversal time \citep{kuai2017examining,williams2020parks,he2020evaluating,salonen2013modelling,scott2008role,neutens2010equity,wang2013planning}. Traversal time is similar to---but distinct from---uncongested travel time. The former merely minimizes the sum of the ratio of street segment length and speed limit, whereas the latter accounts for stops, signals, and turns. Some researchers sidestep the challenges of needing a network model and travel speed data by using secondary traffic analysis zone (TAZ) travel time data aggregated by metropolitan planning agencies \citep[e.g.,][]{grengs2010intermetropolitan,yan2021toward,levine2012does}. These data offer simplicity and real-world empricism, but their coarse-grained zone-to-zone aggregation obfuscates accurate point-to-point travel time predictions.

    Finally, many planning scholars rely instead on the Google Maps API to obtain accurate origin-destination travel times and shortest paths \citep[e.g.,][]{fielbaum2021assessment,costa2021spatial, swayne2021integrating,hu2020estimating,cuervo2022dynamic,chen2020communities,hwang2024measuring}. As mentioned previously, Google travel times represent something of a \enquote{gold standard} in the literature, but come with drawbacks: the algorithms are closed-source and---for large batches of queries, such as for simulating metropolitan-scale trip taking---it can be expensive and require many API queries. Nevertheless, evidence shows Google travel times offer a good approximation of real-world travel times \citep{lin2021impact,fu2023comparative,alsobky2020estimating,wang2011estimating}.

    \subsection{Open Problem}

    In summary, planning researchers today predict travel distances and times with a range of relatively simple techniques that minimize euclidean distance, network distance, street segment traversal time, TAZ-to-TAZ travel times, or Google travel times (for relatively few trips to keep costs down). These techniques are generally inexpensive and easy to implement. In contrast, today's state-of-the-art techniques in the computer science and engineering literatures require expensive data collection, private data, or much more complicated algorithms that planners are rarely trained to implement. In other words, it is expensive and challenging for the average planner to reproduce these state-of-the-art methods.

    A better way would be to seek a middle ground between expensive, complicated, point-to-point empirical travel times and overly-simple naïve predictions. Such free empirical travel time predictions may not reflect congested travel times (which vary drastically throughout the day and week), but would provide a good prediction of minimally-congested travel times. This middle-ground method should meet three criteria. First, it should only use free software and free data, avoiding expensive and massive Google travel time queries or bespoke GPS data collection. Second, it should be a simple and accessible method for urban planners without advanced technical skills. Third, it should provide a substantial improvement in accuracy or spatial resolution over naïve predictions such as minimizing Euclidean distance, network distance, street edge traversal time, and exsiting TAZ-to-TAZ travel times, and approach the accuracy of the state-of-art models. Meeting these three criteria, a middle ground should be inexpensive, easy-to-use, and reasonably accurate.

    \section{Methods}

    This paper proposes a new generalizable method to solve this problem. In short, it collects a small free amount of data from the Google Maps API plus free open data to train a local model to predict accurate minimally-congested travel times using relatively simple standard tools and techniques. Figure~\ref{fig:workflow} shows a detailed workflow.

    \begin{figure*}[hbt!]
        \centering
        \includegraphics[width=1.0\textwidth]{fig_workflow.jpg}
        \caption{Detailed model workflow.}\label{fig:workflow}
    \end{figure*}

    \begin{table*}[hbt!]
        \centering
        \caption{Traffic control elements in the network model, as identified by OSM tags.}\label{tab:node_count}
        \begin{tabular}{lr}
            \toprule
            Traffic control element    & Node count \\ \midrule
            Crossing                   &     21,559 \\
            Stop sign                  &     26,304 \\
            Turning circle             &     17,886 \\
            Traffic signal             &     15,261 \\
            Motorway junction          &      1,436 \\
            Mini roundabout            &         44 \\
            Give way                   &        189 \\
            3-way+ street intersection &    126,982 \\
            Nodes without any tag      &    699,912 \\
            Total nodes                &    782,825 \\ \bottomrule
        \end{tabular}
    \end{table*}

    \subsection{Input Data}

    We define the study area as the convex hull around the intersection of the Los Angeles County boundary and the Los Angeles urban area boundary from the Global Human Settlement Layer's Urban Center Database \citep{florczyk2019description, GHS2019}. This allows us to retain the main urbanized area without adjacent metropolitan areas (such as the Inland Empire). We then model the drivable street network within this study area from OpenStreetMap using the OSMnx software \citep{boeing2017osmnx}, retaining the strongly connected component, to generate a graph with 782,825 nodes. This graph is unsimplified, meaning that nodes represent intersections and dead-ends, as well as curving street line geometry vertices. Table~\ref{tab:node_count} summarizes these nodes' tagged traffic control elements: these data are sparse and most intersections lack traffic control information.

    To generate origin-destination (OD) pairs for training a prediction model, we first over-sample 5,000,000 random node pairs from the street intersections and dead-ends in this graph. We then filter these down to realistic, minimally-congested trip patterns using the most-recently released Uber movement data \citep{ubermovement2020}. These data derive from from Uber trip GPS traces aggregated by tract-to-tract flow and hour of the day. We filter our OD pairs down to those that have a matching real-world trip ($n$ = 1,197,513) that occurred during the 03:00 hour ($n$ = 41,360) to best approximate minimally-congested traffic conditions.

    Finally, we collect \enquote{gold standard} travel times from the Google Maps API (specifically, the Routes API), which has been used in many studies to ground-truth travel times  \citep[e.g.,][]{ludwig2023traffic, hu2020estimating, wang2011estimating, fu2023comparative, delmelle2019travel}. For each OD pair, the API provides the fastest network path, its travel time, and its length. The API allows users to predict travel times for trips departing in the future: the closer the departure time, the more accurate the prediction. Accordingly, we set the departure time to 03:00 and performed the query immediately beforehand (between 02:30--02:40) on 31 January and 1 February 2024. We used Google's \enquote{BEST\_GUESS} traffic-aware model to predict its \enquote{duration\_in\_traffic} travel time. The API was unable to solve 16 OD pairs' routes, resulting in 41,360 OD pairs with a Google travel time, which we used for prediction modeling.

    \begin{figure*}[hbt!]
        \centering
        \includegraphics[width=1.0\textwidth]{fig_turns_definition.jpg}
        \caption{Angular definition of turns in the model.}\label{fig:turns_definition}
    \end{figure*}

    \subsection{Travel Time Prediction}

    Our travel time prediction model has two steps. First, we calculate \enquote{naïve travel times} across a set of trips using open data and open-source software. Second, we train a random forest regression model to predict the Google \enquote{gold standard} travel times from our \enquote{naïve travel times} plus a set of covariates. This subsection details this process.

    To calculate naïve travel times, we solve each OD pair's shortest path using Dijkstra's algorithm to minimize edge traversal time at the speed limit. Then we count the number of traffic control elements (including traffic signals, pedestrian crossings, give ways, and mini roundabouts) and turns (i.e., left, slight left, right, slight right, and u-turn), as shown in Figure~\ref{fig:turns_definition}, along each OD pair's route.

    Then we train a random forest regression model to predict a route's Google travel time from  the naïve travel time plus the count of each type of traffic control element and turn encountered along the naïve shortest path. We also tested other algorithms, including ordinary least squares, gradient boosting regression, AdaBoost regression, and decision tree regression, but settled on random forest regression as its ensemble accuracy substantially outperforms the others.

    We divide our OD pairs into a standard 80/20 training/test split, then tune the hyperparameters through a randomized grid search with 5-fold cross-validation by calculating their mean absolute error (MAE), defined in Equation~\ref{eq:mae}:

    \begin{equation}
        \label{eq:mae}
        \text{MAE} = \frac{\sum^{n}_{i=1} \left|{t_i - \hat{t_i}}\right|}{n}
    \end{equation}

    where $n$ is the total number of trips, $t_i$ is trip $i$'s observed travel time, and $\hat{t_i}$ is trip $i$'s predicted travel time.

    \subsection{Validation}

    We validate our out-of-sample predictions against the \enquote{gold standard} Google travel times using five measures. First, we calculate the MAPE, following Equation~\ref{eq:mape}. Second, we calculate the MAE, following Equation~\ref{eq:mae}, to measure the average absolute difference in seconds between our prediction and the reference Google value. Third, we perform difference-in-means ($\delta$) $t$-tests to determine whether our predicted travel times are statistically significantly different (at the $p<=0.05$ significance level) from the reference Google values. Despite the large sample size, we expect insignificant $t$-statistics if the predicted travel times' distribution approximates the Google travel times' distribution. Fourth, we calculate the standard $R^2$ to measure our model's success in explaining travel time variation. Fifth and finally, we calculate the average of the pairwise ratios (APR) of our predicted travel time and the Google travel time to measure the average over- or under-prediction of the trips' travel times, as defined in Equation~\ref{eq:apr}:

    \begin{equation}
        \label{eq:apr}
        \text{APR} = \frac{\sum^{n}_{i=1}\frac{t_i}{\hat{t_i}}}{n}
    \end{equation}

    where $n$ is the total trip count, $t_i$ is trip $i$'s Google travel time, and $\hat{t_i}$ is our predicted travel time for trip $i$.

    \section{Results}

    \subsection{Optimized Hyperparameters}

    Our final optimized model hyperparameterization uses 600 decision trees, random sampling with replacement in each decision tree, a maximum decision tree depth of 10, a non-limited maximum number of input explanatory features, a requirement of at least two samples to split at a decision node and one sample at a leaf node, and equal sample weighting. Finally, we check our trained model for overfitting by conducting 5-fold cross-validation across the whole sample using tuned hyperparameters: the resulting five MAE values (a, b, c, d, e) are all similar to each other, which indicates no overfitting problem as the hyperparameters are not overly specific to the training set.

    \subsection{Model Performance and Validation}

    Table~\ref{tab:validation_results} details our travel time prediction model's out-of-sample improvement over the initial naïve travel time calculation. Our model achieves a MAPE of 8.4\%---in line with the \textasciitilde10\% MAPEs seen in the state-of-the-art travel time prediction literature, but without their extensive and expensive input data requirements. In comparison, our initial naïve travel time calculation yields a much worse MAPE of 21.1\%. Whereas the naïve travel time calculation's MAE is 183.5 seconds from the Google travel time, our predicted travel times' MAE is just 75.2 seconds---an improvement by a factor of 2.4.

    \begin{table*}[tb!]
        \centering
        \caption{Results of out-of-sample (testing set) prediction of 1) our prediction model and 2) the initial naïve travel time calculation, validated against the corresponding Google travel times. The five validation measures are 1) MAPE, 2) MAE, 3) difference-in-means ($\delta$) and corresponding $t$-test $p$-value, 4) APR, and 5) $R^2$. See methods for definitions.}\label{tab:validation_results}
        \begin{tabular}{lrr}
            \toprule
            & Prediction model & Naïve calculation \\
            \midrule
            MAPE (\%)                     &             8.43 &             21.13 \\
            MAE (seconds)                 &            75.23 &            183.55 \\
            $\delta$ (seconds)            &             0.38 &           -182.71 \\
            $p$-value                     &             0.75 &             <0.01 \\
            APR                           &             1.01 &              0.79 \\
            $R^2$                         &             0.93 &              0.74 \\
            \midrule
            $n$                           &           8,272 &            8,272 \\
            \bottomrule
        \end{tabular}
    \end{table*}

    The $t$-test reveals significant differences between the initial naïve travel time calculations and the Google travel times: the $\delta$ of -182.7 seconds corresponds to $p<0.01$. In other words, the naïve travel time under-predicts Google by over 3 minutes. However, the $t$-test reveals insignificant differences between our model's out-of-sample predictions and the Google travel times: the $\delta$ of 0.4 seconds corresponds to a $p$ value of 0.75. That is, our prediction model over-predicts by less than half a second on average, which is statistically insignificantly different from zero. If we compare the aforementioned MAEs to these differences-in-means, we can see that the initial naïve travel time calculations significantly and consistently under-predict travel time, but our prediction model shows no such directional bias: its random error balances out between (smaller absolute) over-and under-prediction.

    The APR offers another lens on this finding. On average across the OD pairs, the initial naïve travel time calculation under-predicts Google travel time by 21\%, but our prediction model over-predicts it by just 1\%. Our model also explains more of the variation in travel time: its $R^2$ of 0.93 is substantially higher than the 0.74 $R^2$ of the initial naïve travel time calculation. In summary, each of our five validation measures demonstrates that our prediction model drastically improves on naïve travel time calculations---without needing extensive, expensive input data or advanced deep learning software---and offers travel time predictions much closer to those of Google's \enquote{gold standard} predictions.

    \section{Discussion}
    Our study shows that our proposed method successfully meets its three intended criteria for middle-ground travel prediction: free, easy implementation, and high prediction accuracy. First, the model is built entirely on free and open-source data and software. It uses OpenStreetMap (OSM) for the street network and only a small amount of Google travel time data within the monthly free quota. In addition, its modeling process relies on open-source Python packages, including OSMnx for network routing and scikit-learn for training the random forest regression model. Second, this method is simple and easy to use. Rather than requiring sophisticated deep-learning techniques in the computer science literature, the proposed method only uses an out-of-the-box random forest regression model that can be implemented by anyone with basic experience in data science. Third, in terms of accuracy, our model's MAPE of 8.43\% substantially improves the MAPE of 21.13\% in the naïve model and falls within the MAPE range of 3\% to 17\% in the state-of-the-art models that we found in the recent literature. In summary, these results confirm that we proposed a free, easy-to-use, and accurate method for predicting minimally-congested travel time. It provides a middle-ground solution to improve travel time prediction accuracy without relying on expensive datasets and complicated methods.

    Importantly, these strengths are especially valuable for the audience that need them most---the urban planners. Although our proposed method does not aim to outperform the cutting-edge models, it provides an improved middle-ground solution for urban planners who lack access to expensive datasets and technical models typical in the computer science literature. Instead, they often rely on simple but less accurate methods, such as minimizing Euclidean distance, network distance, or edge traversal time based on the speed limit. These naïve methods, while easy to compute, systematically under-predict real-world travel times because they ignore the delays caused by turns and traffic controls. Our model addresses this problem by incorporating these features into the random forest regression model, which is much easier to implement than sophisticated deep learning models. In other words, our method does not replace the state-of-the-art methods for those who can implement them, but it makes meaningful progress over the naïve methods used by most urban planners.

    Building on these strengths, this study offers three contributions to planning research and practice. First, it provides a practical solution to bridge the methodological gap between overly complex and overly simple travel time predictions. On the one hand, accurate predictions of traffic-aware travel time require expensive, extensive, and often proprietary GPS data, which is often inaccessible to many urban planners. On the other hand, planners often use simple methods that under-predict real-world travel times. Our method provides a viable middle ground: a minimally congested travel time prediction method that is open, free, and easy-to-compute. Second, in addition to methodological accessibility, this model supports high-resolution, point-to-point travel time predictions across a large metropolitan area. Our case study is based on the populated area of Los Angeles County, demonstrating its applicability for metropolitan-scale travel time prediction. Third, the model is lightweight and efficient. It requires relatively short runtimes while maintaining prediction accuracy comparable to complex models. These contributions are important in urban planning practices because travel time can shift downstream planning processes based on accessibility, travel mode choice, and land use analysis. Inaccurate travel times can skew the results and cause misguided planning interventions. For example, based on under-predicted travel times, planners may falsely interpret high accessibility levels for marginalized neighborhoods far from employment centers. This could mask the real-world accessibility disadvantages and lead to underinvestment in these communities. Therefore, our proposed travel prediction method helps reduce transportation modeling bias and supports evidence-based planning interventions.

    Despite these contributions, our study has limitations that point future research directions. First, while used by billions of people around the world as the \enquote{best available} source of travel time prediction, Google travel time remains a black-box model based on users' GPS data. Future studies could use real-world observations, such as travel time from GPS traces, to train and validate the model. Second, our method is trained to predict minimally-congested travel time at 03:00 a.m. While it significantly improves upon naive travel time predictions, it does not reflect peak-hour traffic delays. Future studies could expand the model to different times of day to test its applicability. Additionally, our method predicts travel time but does not estimate the actual routes drivers may take. Given these limitations, it is not designed for cities with resources to develop sophisticated traffic models and extensive real-time urban informatics dashboards on travel demand and traffic flows. Rather, it serves urban planners in research and practice who lack access to such systems and need a low-cost technique for predicting better travel times than naïve methods. Third, our case study looked at the populated area in Los Angeles County, where the information in OpenStreetMap is relatively complete. To test the generalizability of this method, further studies should examine its performance in cities with sparser or lower-quality open data. Taken together, these limitations open opportunities to further improve travel time predictions across diverse geographical and temporal settings in the future.

    \section{Conclusion}

    Travel time prediction is central to questions of urban accessibility, mode and route choice, and individuals’ location decisions. True congested travel time vary drastically throughout the day and require large volumes of real-time or proprietary data, as well as complex algorithms, to model. These technical challenges and costs present a hurdle for many urban planners, who often turn to simpler models with lower and often free data and computing requirements. The traditional naïve methods are easy to implement, but produce inaccurate predictions.

    This study introduces a middle-ground method to simply but accurately predict minimally-congested travel time using free data. It improves on traditional, common, naïve predictions by incorporating a small one-off collection of free but proprietary high-quality Google data on travel times. Also of note is the sparseness of our open data: OpenStreetMap contains little information on traffic controls, as only around 28\% of the intersections in the study area have traffic control information. Even so, our prediction model using random forest regression still predicts Google travel times with high accuracy. This offers planning practitioners and scholars a better method to predict travel times for free with much better accuracy than traditional naïve methods offered.

    How can it be reused by others?
    Can it be extended to other times of day?
    Can it be extended to other study sites?

    \section*{Acknowledgments}

    The authors wish to thank Youngseo Kweon and Jaehyun Ha for additional research assistance.

    % print the footnotes as endnotes, if any exist
    \IfFileExists{\jobname.ent}{\theendnotes}{}

    % print the bibliography
    \setlength{\bibsep}{0.00cm plus 0.05cm} % no space between items
    \bibliographystyle{apalike}
    \bibliography{references}

\end{document}

### 2025-04-15 scripts/04-random_forest.py
Fixed a minor bug and assigned random seed for consistency

@@ -14,6 +14,7 @@ from sklearn.model_selection import (
     RandomizedSearchCV,
     cross_val_score,
     train_test_split,
+    KFold
 )
 
 
@@ -64,11 +65,12 @@ def rf_model(df: pd.DataFrame) -> tuple[pd.DataFrame, pd.DataFrame, np.ndarray, 
     rf.fit(x, y)
     # Predict the result
     test["rf_predict_default"] = rf.predict(x_test)
+    kf = KFold(n_splits=5, shuffle=True, random_state=123)
     base_cross_val_score = cross_val_score(
         rf,
         df[list_of_features],
         df["duration"],
-        cv=5,
+        cv=kf,
         scoring="neg_mean_absolute_error",
     )
 
@@ -115,13 +117,13 @@ def rf_model(df: pd.DataFrame) -> tuple[pd.DataFrame, pd.DataFrame, np.ndarray, 
         best_random,
         df[list_of_features],
         df["duration"],
-        cv=5,
+        cv=kf,
         scoring="neg_mean_absolute_error",
     )
     test.to_csv(constants.PREDICTION_RESULT_FILE_PATH)
 
     # Write the results to a text file
-    with Path.open(constants.NETWORK_ROUTING_EVALUATION_RESULT_FILE_PATH, "w") as file:
+    with Path.open(constants.CROSS_VALIDATION_RESULT_FILE_PATH, "w") as file:
         file.write(f"Base Cross-Validation Score: {base_cross_val_score}\n")
         file.write(f"Tuned Cross-Validation Score: {tuned_cross_val_score}\n")
 

### 2025-04-15 scripts/04-random_forest.py
minor format fix

@@ -10,12 +10,7 @@ import pandas as pd
 from scipy import stats
 from sklearn.ensemble import RandomForestRegressor
 from sklearn.metrics import r2_score
-from sklearn.model_selection import (
-    RandomizedSearchCV,
-    cross_val_score,
-    train_test_split,
-    KFold
-)
+from sklearn.model_selection import KFold, RandomizedSearchCV, cross_val_score, train_test_split
 
 
 def read_result() -> pd.DataFrame:

### 2025-04-17 paper/main.tex

@@
 \affil[]{Redacted for review}%{\myaffiliation}
-    \date{ }%{\paperdate}
+\date{}%{\paperdate}
 
@@
 These obstacles make it difficult for researchers, practitioners, and community advocates to develop an evidence base for more sustainable transportation and equitable accessibility.
-This paper develops a free, open-source minimally-congested driving time prediction model with minimal cost, data, and computational requirements.
+This article develops a free, open-source minimally-congested driving time prediction model with minimal cost, data, and computational requirements.
 It trains and tests this model using the Los Angeles, California urban area as a case study by calculating naïve travel times then developing a predictive random forest regression model of Google Maps' \enquote{gold standard} travel times using those naïve times plus turns and traffic controls along the naïve routes.
@@
 This is usually not a problem for well-resourced organizations with highly-skilled labor.
-However, technical expertise, complicated data and software requirements, and high costs pose insurmountable obstacles for many urban planning researchers, practitioners, and community advocates.
+However, technical expertise, complex data and software requirements, and high costs pose insurmountable obstacles for many urban planning researchers, practitioners, and community advocates.
 
-    Although typical \enquote{naïve} travel time prediction methods that minimize travel distance or street segment traversal time provide easy solutions, they ignore delays from turns and traffic controls and thus usually under-predicts real-world travel time.
+Although typical \enquote{naïve} travel time prediction methods (which minimize travel distance or street segment traversal time) provide easy solutions, they ignore delays from turns and traffic controls and thus usually under-predict real-world driving travel time.
 However, minimally-congested travel time prediction offers a middle ground of more-accurate predictions than naïve travel times do---yet much lower data and computing requirements than traffic-aware and congested travel time prediction.
-Focusing on driving, this study contributes a simple, accurate, and free method of minimally-congested travel time prediction.
+
+To address this need for a middle ground, this article contributes a simple, accurate, and free method of minimally-congested driving travel time prediction.
 Our results show a substantial improvement over traditional, naïve methods of travel time prediction common in the urban planning literature.
-As we shall show in this paper, across the Los Angeles urban area, our average trip's travel time prediction differs on average from Google's \enquote{gold standard} prediction by only 0.38 seconds.
+As we shall show in this article, across the Los Angeles urban area, our average trip's travel time prediction differs on average from Google's \enquote{gold standard} prediction by only 0.38 seconds.
 In contrast, simple---but common---naïve travel time prediction minimizing max-speed edge traversal times under-predicts Google's travel times by 182.71 seconds, on average.
 
-    The rest of this paper is organized as follows.
+The rest of this article is organized as follows.
 First it reviews the state of the art in travel time prediction and then considers the less-sophisticated methods that are standard in the current planning literature.
@@
 
-    Modern travel time prediction techniques tend to fall into one of two camps: 1) sophisticated, accurate, but difficult to execute; or 2) naïve, less-accurate, but simple to execute.
+Modern travel time prediction techniques tend to fall into one of two camps: (1) sophisticated, accurate, but difficult to execute; or (2) naïve, less-accurate, but simple to execute.
 The former is the state-of-the-art and has been the subject of much work in the recent computer science and engineering literatures.
@@
 
-    These advanced travel time prediction models require massive data input as they are usually trained on millions or billions of disaggregate empirical travel time records.
+These advanced travel time prediction models require massive data inputs as they are usually trained on millions or billions of disaggregate empirical travel time records.
 For example, \citet{hou2018network} use 1.5 billion GPS-based travel time records in deep learning models to predict travel time in 13 road segments in St.
@@
 Using more than 100,000 trips from Windows phone GPS data, \citet{woodard2017predicting}'s technique achieves a MAPE of 10.1\% when predicting travel times.
-\citet{pamula2023estimation} use three months of video sensing data at a 5-minute temporal resolution in a deep learning model to predict travel times in Poland, yielding a 6.8\% MAPE.
-Other studies use graph neural network models to predict travel times.
+\citet{pamula2023estimation} use three months of video sensing data at a 5-minute temporal resolution in a deep learning model to predict travel times in Poland, yielding a 6.8\% MAPE.\@ Other studies use graph neural network models to predict travel times.
 For example, \citet{wang2023dynamic} use data from 100,000 Chinese ride-hailing trips in graph neural network and recurrent neural network models, yielding a MAPE of 15.4\%.
@@
 
-    Central to all of the preceding travel time prediction research is the concept of \textit{path solving}: identifying the shortest path between an origin and destination in a network model.
+Central to all of the preceding travel time prediction research is the concept of \textit{path solving}: identifying the shortest path between an origin and destination in a spatial network model.
 In this case, the shortest path is defined as the path that minimizes travel time and could be realized methodologically in multiple ways.
@@
 For example, many accessibility studies employ merely a (very) rough proxy for travel time by minimizing euclidean distance traveled instead \citep[e.g.,][]{macfarlane2021modeling,pearsall2020locating}.
-This offers the substantial benefit of simplicity (no network model, travel time date, or routing algorithms are needed) but offers a poor estimate of access in terms of travel time.
+This offers the substantial benefit of simplicity (no network model, travel time data, or routing algorithms are needed) but offers a poor estimate of access in terms of travel time.
 
-    Many other studies improve on this by instead solving shortest paths by minimizing network distance traveled.
+Other studies improve on this by instead solving shortest paths by minimizing network distance traveled.
 These studies often use data and tools from Esri, OpenStreetMap, or the US Census Bureau to measure network distances between origins and destinations \citep[e.g.,][]{mckenzie2020urban, jiao2021measuring, nicoletti2023disadvantaged, logan2019evaluating, tsou2005accessibility}.
@@
 Finally, many planning scholars rely instead on the Google Maps API to obtain accurate origin-destination travel times and shortest paths \citep[e.g.,][]{fielbaum2021assessment,costa2021spatial, swayne2021integrating,hu2020estimating,cuervo2022dynamic,chen2020communities,hwang2024measuring}.
-As mentioned previously, Google travel times represent something of a \enquote{gold standard} in the literature, but come with drawbacks: the algorithms are closed-source and---for large batches of queries, such as for simulating metropolitan-scale trip taking---it can be expensive and require many API queries.
+As mentioned previously, Google travel times represent something of a \enquote{gold standard} in the literature, but come with drawbacks: the algorithms are closed-source and---for large batches of queries, such as for simulating metropolitan-scale trip taking---it can become prohibitively expensive and require many API queries.
 Nevertheless, evidence shows Google travel times offer a good approximation of real-world travel times \citep{lin2021impact,fu2023comparative,alsobky2020estimating,wang2011estimating}.
@@
 
-    This paper proposes a new generalizable method to solve this problem.
+This article proposes a new generalizable method to solve this problem.
 In short, it collects a small free amount of data from the Google Maps API plus free open data to train a local model to predict accurate minimally-congested travel times using relatively simple standard tools and techniques.
@@
 This allows us to retain the main urbanized area without adjacent metropolitan areas (such as the Inland Empire).
-We then model the drivable street network within this study area from OpenStreetMap using the OSMnx software \citep{boeing2017osmnx}, retaining the strongly connected component, to generate a graph with 782,825 nodes.
+We then model the drivable street network within this study area from OpenStreetMap using the OSMnx software \citep{boeing_modeling_2025}, retaining the strongly connected component, to generate a graph with 782,825 nodes.
 This graph is unsimplified, meaning that nodes represent intersections and dead-ends, as well as curving street line geometry vertices.
@@
 Our final optimized model hyperparameterization uses 600 decision trees, random sampling with replacement in each decision tree, a maximum decision tree depth of 10, a non-limited maximum number of input explanatory features, a requirement of at least two samples to split at a decision node and one sample at a leaf node, and equal sample weighting.
-Finally, we check our trained model for overfitting by conducting 5-fold cross-validation across the whole sample using tuned hyperparameters: the resulting five MAE values (a, b, c, d, e) are all similar to each other, which indicates no overfitting problem as the hyperparameters are not overly specific to the training set.
+Finally, we check our trained model for overfitting by conducting 5-fold cross-validation across the whole sample using tuned hyperparameters: the resulting five MAE values (75.3, 74.5, 73.5, 73.3, 74.7) are all similar to each other, which indicates no overfitting problem as the hyperparameters are not overly specific to the training set.
 
@@
 Table~\ref{tab:validation_results} details our travel time prediction model's out-of-sample improvement over the initial naïve travel time calculation.
-Our model achieves a MAPE of 8.4\%---in line with the \textasciitilde10\% MAPEs seen in the state-of-the-art travel time prediction literature, but without their extensive and expensive input data requirements.
+Our model achieves a MAPE of 8.4\%, in line with the approximately 10\% MAPEs seen in the state-of-the-art travel time prediction literature, but without their extensive and expensive input data requirements.
 In comparison, our initial naïve travel time calculation yields a much worse MAPE of 21.1\%.
@@
     \centering
-        \caption{Results of out-of-sample (testing set) prediction of 1) our prediction model and 2) the initial naïve travel time calculation, validated against the corresponding Google travel times.
-The five validation measures are 1) MAPE, 2) MAE, 3) difference-in-means ($\delta$) and corresponding $t$-test $p$-value, 4) APR, and 5) $R^2$.
+    \caption{Results of out-of-sample (testing set) prediction of (1) our prediction model and (2) the initial naïve travel time calculation, validated against the corresponding Google travel times.
+The five validation measures are (1) MAPE, (2) MAE, (3) difference-in-means ($\delta$) and corresponding $t$-test $p$-value, (4) APR, and (5) $R^2$.
 See methods for definitions.}\label{tab:validation_results}
@@
 \section{Discussion}
-    Our study shows that our proposed method successfully meets its three intended criteria for middle-ground travel prediction: free, easy implementation, and high prediction accuracy.
-First, the model is built entirely on free and open-source data and software.
-It uses OpenStreetMap (OSM) for the street network and only a small amount of Google travel time data within the monthly free quota.
-In addition, its modeling process relies on open-source Python packages, including OSMnx for network routing and scikit-learn for training the random forest regression model.
-Second, this method is simple and easy to use.
-Rather than requiring sophisticated deep-learning techniques in the computer science literature, the proposed method only uses an out-of-the-box random forest regression model that can be implemented by anyone with basic experience in data science.
-Third, in terms of accuracy, our model's MAPE of 8.43\% substantially improves the MAPE of 21.13\% in the naïve model and falls within the MAPE range of 3\% to 17\% in the state-of-the-art models that we found in the recent literature.
-In summary, these results confirm that we proposed a free, easy-to-use, and accurate method for predicting minimally-congested travel time.
-It provides a middle-ground solution to improve travel time prediction accuracy without relying on expensive datasets and complicated methods.
 
-    Importantly, these strengths are especially valuable for the audience that need them most---the urban planners.
-Although our proposed method does not aim to outperform the cutting-edge models, it provides an improved middle-ground solution for urban planners who lack access to expensive datasets and technical models typical in the computer science literature.
-Instead, they often rely on simple but less accurate methods, such as minimizing Euclidean distance, network distance, or edge traversal time based on the speed limit.
-These naïve methods, while easy to compute, systematically under-predict real-world travel times because they ignore the delays caused by turns and traffic controls.
-Our model addresses this problem by incorporating these features into the random forest regression model, which is much easier to implement than sophisticated deep learning models.
-In other words, our method does not replace the state-of-the-art methods for those who can implement them, but it makes meaningful progress over the naïve methods used by most urban planners.
+Re-state the problem.
+This study set out to investigate if a middle-ground travel time prediction model was possible.
+It identified three criteria for success: (1) it should use free software and data and avoid bespoke sensor data collection; (2) it should be easy and accessible to use; (3) it should offer better accuracy than naïve predictions (e.g., minimizing Euclidean distance, network distance, street edge traversal time) and better spatial resolution than zone-to-zone travel times.
 
-    Building on these strengths, this study offers three contributions to planning research and practice.
-First, it provides a practical solution to bridge the methodological gap between overly complex and overly simple travel time predictions.
-On the one hand, accurate predictions of traffic-aware travel time require expensive, extensive, and often proprietary GPS data, which is often inaccessible to many urban planners.
-On the other hand, planners often use simple methods that under-predict real-world travel times.
-Our method provides a viable middle ground: a minimally congested travel time prediction method that is open, free, and easy-to-compute.
-Second, in addition to methodological accessibility, this model supports high-resolution, point-to-point travel time predictions across a large metropolitan area.
-Our case study is based on the populated area of Los Angeles County, demonstrating its applicability for metropolitan-scale travel time prediction.
-Third, the model is lightweight and efficient.
-It requires relatively short runtimes while maintaining prediction accuracy comparable to complex models.
-These contributions are important in urban planning practices because travel time can shift downstream planning processes based on accessibility, travel mode choice, and land use analysis.
-Inaccurate travel times can skew the results and cause misguided planning interventions.
-For example, based on under-predicted travel times, planners may falsely interpret high accessibility levels for marginalized neighborhoods far from employment centers.
-This could mask the real-world accessibility disadvantages and lead to underinvestment in these communities.
-Therefore, our proposed travel prediction method helps reduce transportation modeling bias and supports evidence-based planning interventions.
+Our results show that our proposed method satisfies these criteria to offer an important middle-ground contribution.
+First, it relies only on open-source Python software, OpenStreetMap data, and a small free amount of Google travel time training data.
+Second, our model is easy to use: it does not require a complex computing environment suited for deep learning's extensive data and processing requirements.
+Rather, our model uses out-of-the-box tools that can be set up in seconds on a standard consumer-grade computer by anyone familiar with Python, the world's most popular programming language.
+Third, our model is accurate: its MAPE (8.4\%) is much lower than that of the naïve model and in line with state-of-the-art models (\textasciitilde3--17\%) with extensive data and computing requirements.
 
-    Despite these contributions, our study has limitations that point future research directions.
-First, while used by billions of people around the world as the \enquote{best available} source of travel time prediction, Google travel time remains a black-box model based on users' GPS data.
-Future studies could use real-world observations, such as travel time from GPS traces, to train and validate the model.
-Second, our method is trained to predict minimally-congested travel time at 03:00 a.m.
-While it significantly improves upon naive travel time predictions, it does not reflect peak-hour traffic delays.
-Future studies could expand the model to different times of day to test its applicability.
-Additionally, our method predicts travel time but does not estimate the actual routes drivers may take.
-Given these limitations, it is not designed for cities with resources to develop sophisticated traffic models and extensive real-time urban informatics dashboards on travel demand and traffic flows.
-Rather, it serves urban planners in research and practice who lack access to such systems and need a low-cost technique for predicting better travel times than naïve methods.
-Third, our case study looked at the populated area in Los Angeles County, where the information in OpenStreetMap is relatively complete.
-To test the generalizability of this method, further studies should examine its performance in cities with sparser or lower-quality open data.
-Taken together, these limitations open opportunities to further improve travel time predictions across diverse geographical and temporal settings in the future.
+This middle ground is not necessarily for computer scientists, high-end labs with top-of-the-line computing hardware, or well-resourced agencies with extensive data budgets.
+Rather, it empowers the scholars and practitioners who need it the most---those working on the frontlines of urban planning and policymaking.
+These scholars and practitioners rely on simple but inaccurate routing models---such as minimizing Euclidean distance, network distance, or edge traversal time---when they lack the technical expertise, computing equipment, or extensive data needed to implement the literature's cutting-edge models.
+However, this study demonstrated how these naïve methods systematically under-predict real-world travel times.
+Our model addresses this problem by incorporating sparse open data on traffic control elements and turns into the prediction and achieves an accuracy in line with state-of-the-art methods---but without their data needs or computing requirements.
+In other words, our model does not replace the state-of-the-art methods for those with extensive resources available to implement them, but rather offers far more accurate predictions for urban planners who would otherwise fall back on simpler naïve models.
+Inaccurate travel times skew empirical understandings and mislead planning interventions.
+Our model represents an important contribution for urban planning scholarship and evidence-based practitioner interventions.
+
+As demonstrated, this model supports high-resolution point-to-point travel time predictions across a large metropolitan area.
+Our implementation focused on Los Angeles but is not inherently tied to it, and future work should emphasize its generalizability in cities with potentially sparser or lower-quality open data.
+We trained this model as a proof-of-concept for minimally-congested times of day, but the same concept can be extended with similar training data for any other time of day.
+Future research should also expand the kinds of training and validation data.
+Although used by billions of people around the world as the public's \enquote{best available} source of travel time prediction, Google travel time remains a black box based on users' GPS data.
+Future work can use other empirical travel time data to further train and validate the model.
 
@@
 
-    Travel time prediction is central to questions of urban accessibility, mode and route choice, and individuals’ location decisions.
+Travel time prediction is central to questions of urban accessibility, mode and route choice, and individuals' location decisions.
 True congested travel time vary drastically throughout the day and require large volumes of real-time or proprietary data, as well as complex algorithms, to model.
@@
 
-    This study introduces a middle-ground method to simply but accurately predict minimally-congested travel time using free data.
-It improves on traditional, common, naïve predictions by incorporating a small one-off collection of free but proprietary high-quality Google data on travel times.
+This article introduced a middle-ground method to simply but accurately predict minimally-congested travel time using free data.
+It improves on traditional, common, naïve predictions by training on a small one-off collection of free but proprietary high-quality Google data on travel times.
 Also of note is the sparseness of our open data: OpenStreetMap contains little information on traffic controls, as only around 28\% of the intersections in the study area have traffic control information.
@@
 This offers planning practitioners and scholars a better method to predict travel times for free with much better accuracy than traditional naïve methods offered.
-
-    How can it be reused by others?
-    Can it be extended to other times of day?
-    Can it be extended to other study sites?
 

### 2025-04-30 paper/main.tex
update paper

@@
 
-% import encoding and font packages for pdflatex, in order
-\usepackage[T1]{fontenc} % output T1 font encoding (8-bit) so accented characters are a single glyph
+% load encoding and font packages for pdflatex, in order
+\usepackage[T1]{fontenc}    % output 8-bit encoded fonts
 \usepackage[utf8]{inputenc} % allow input of utf-8 encoded characters
@@
 
-% import language, regionalization, microtype, and setspace in order
+% load babel, csquotes, and microtype in order
+\usepackage[USenglish]{babel} % auto-regionalize hyphens, quote marks, etc
 \usepackage[strict,autostyle]{csquotes} % smart and nestable quote marks
-\usepackage[USenglish]{babel} % automatically regionalize hyphens, quote marks, etc
-\usepackage{microtype} % improves text appearance with kerning, etc
-\usepackage{setspace} % configure spacing between lines
-\usepackage{mathtools} % use package math tool
+\usepackage[babel=true]{microtype} % enable micro-typographic adjustments
+
 % import everything else
-\usepackage{abstract} % allow full-page title/abstract in twocolumn mode
-\usepackage{amsmath}
+\usepackage{amsmath} % additional mathematical typesetting features
 \usepackage{authblk} % footnote-style author/affiliation info
@@
 \usepackage{graphicx} % better inclusion of graphics
-\usepackage{hyperref} % hypertext in document
 \usepackage{natbib} % author-year citations w/ bibtex, including textual and parenthetical
 \usepackage{rotating} % rotate wide tables or figures on a page to make them landscape
+\usepackage{setspace} % configure spacing between lines
 \usepackage{titlesec} % custom section and subsection heading
 \usepackage{url} % make nice line-breakble urls
-\usepackage{makecell}
+
+% load hyperref/orcidlink last for compatibility
+\usepackage{hyperref} % enable hyperlinks and pdf metadata
+\usepackage{orcidlink} % provide orcid logo and link
 
@@
 % location of figure files, via graphicx package
-\graphicspath{{./figures/}}
+\graphicspath{{.}}
 
@@
 \affil[]{Redacted for review}%{\myaffiliation}
-\date{}%{\paperdate}
+\date{}%\paperdate}
 
@@
 Travel time prediction is central to accessibility analyses, sustainable transportation infrastructure provision, and active transportation interventions.
-However, calculating accurate travel times, especially for driving, requires either extensive technical capacities and bespoke data, or sources like the Google Maps API that quickly becomes prohibitively expensive to analyze thousands or millions of trips necessary for metropolitan-scale planning.
+However, calculating accurate travel times, especially for driving, requires either extensive technical capacity and bespoke data, or sources like the Google Maps API that quickly become prohibitively expensive to analyze thousands or millions of trips necessary for metropolitan-scale planning.
 These obstacles make it difficult for researchers, practitioners, and community advocates to develop an evidence base for more sustainable transportation and equitable accessibility.
-This article develops a free, open-source minimally-congested driving time prediction model with minimal cost, data, and computational requirements.
-It trains and tests this model using the Los Angeles, California urban area as a case study by calculating naïve travel times then developing a predictive random forest regression model of Google Maps' \enquote{gold standard} travel times using those naïve times plus turns and traffic controls along the naïve routes.
-Validation reveals that this method yields accurate, free, and simple travel time predictions.
+This article introduces a free, open-source minimally-congested driving time prediction model with minimal cost, data, and computational requirements.
+It trains and tests this model using the Los Angeles, California urban area as a case study by calculating naïve travel times from open data then developing a random forest model to predict true travel times as a function of those naïve times plus open data on turns and traffic controls.
+Validation shows that this method produces accurate, free, and simple travel time predictions, offering a superior middle-ground technique for urban planners.
 
@@
 Travel time prediction underpins many transport planning processes and research problems.
-It is key to understanding urban accessibility, transport mode and route choice, and location decisions.
-Predicting congested travel times (i.e., when traffic congestion impedes flows) requires large volumes of real-time data on traffic conditions.
-Such data are proprietary, expensive, and beyond the reach of many planning practitioners and scholars.
-This is usually not a problem for well-resourced organizations with highly-skilled labor.
-However, technical expertise, complex data and software requirements, and high costs pose insurmountable obstacles for many urban planning researchers, practitioners, and community advocates.
+It is key to understanding urban accessibility, transport mode and route choice, and location decisions \citep{zhang2015gradient,jenelius2013travel}.
+Predicting congested travel times (i.e., when traffic congestion impedes flows) requires large volumes of real-time data on traffic conditions \citep[e.g.,][]{hou2018network}.
+Such data are proprietary, expensive, and beyond the reach of many planning practitioners and scholars \citep{carrion2012value,wang2011estimating}.
+This is usually not a problem for well-resourced organizations with highly-skilled labor---however, technical expertise, complex data and software requirements, and high costs pose insurmountable obstacles for many urban planning researchers, practitioners, and community advocates \citep{giles2022creating}.
 
-Although typical \enquote{naïve} travel time prediction methods (which minimize travel distance or street segment traversal time) provide easy solutions, they ignore delays from turns and traffic controls and thus usually under-predict real-world driving travel time.
-However, minimally-congested travel time prediction offers a middle ground of more-accurate predictions than naïve travel times do---yet much lower data and computing requirements than traffic-aware and congested travel time prediction.
+Although typical \enquote{naïve} travel time prediction methods (which minimize travel distance or street segment traversal time) provide easy solutions, they ignore delays from turns and traffic controls and thus usually under-predict real-world driving travel time \citep{ludwig2023traffic,salonen2013modelling,yiannakoulias2013estimating}.
+However, as we will argue, minimally-congested travel time prediction offers a better middle ground of more-accurate predictions than naïve travel times do---yet much lower data and computing requirements than real-time traffic-aware congested travel time prediction.
 
-To address this need for a middle ground, this article contributes a simple, accurate, and free method of minimally-congested driving travel time prediction.
+To address this need for a middle ground, this article contributes a novel accurate, easy-to-use, and free method of minimally-congested driving travel time prediction.
 Our results show a substantial improvement over traditional, naïve methods of travel time prediction common in the urban planning literature.
-As we shall show in this article, across the Los Angeles urban area, our average trip's travel time prediction differs on average from Google's \enquote{gold standard} prediction by only 0.38 seconds.
-In contrast, simple---but common---naïve travel time prediction minimizing max-speed edge traversal times under-predicts Google's travel times by 182.71 seconds, on average.
+As we shall show in this article, across the Los Angeles urban area, our average trip's travel time prediction differs from Google's \enquote{gold standard} prediction by only 0.45 seconds.
+In contrast, simple---but common---naïve travel time prediction minimizing max-speed edge traversal times under-predicts by 183.55 seconds on average.
 
@@
 Recent studies in the computer science and engineering literatures propose a variety of prediction techniques with heavy computing and data requirements.
-They usually measure their prediction accuracy by mean absolute percentage error (MAPE), defined in Equation~\ref{eq:mape}:
+They usually measure their prediction accuracy by mean absolute percentage error (MAPE), defined in \autoref{eq:mape}:
 
@@
 As mentioned previously, Google travel times represent something of a \enquote{gold standard} in the literature, but come with drawbacks: the algorithms are closed-source and---for large batches of queries, such as for simulating metropolitan-scale trip taking---it can become prohibitively expensive and require many API queries.
-Nevertheless, evidence shows Google travel times offer a good approximation of real-world travel times \citep{lin2021impact,fu2023comparative,alsobky2020estimating,wang2011estimating}.
+Nevertheless, evidence shows Google travel times offer a good proxy for observed real-world travel times \citep{lin2021impact,fu2023comparative,alsobky2020estimating,wang2011estimating}.
 
@@
 This middle-ground method should meet three criteria.
-First, it should only use free software and free data, avoiding expensive and massive Google travel time queries or bespoke GPS data collection.
+First, it should only use free software and free data, avoiding expensive API queries or bespoke GPS data collection.
 Second, it should be a simple and accessible method for urban planners without advanced technical skills.
@@
 
-This article proposes a new generalizable method to solve this problem.
-In short, it collects a small free amount of data from the Google Maps API plus free open data to train a local model to predict accurate minimally-congested travel times using relatively simple standard tools and techniques.
-Figure~\ref{fig:workflow} shows a detailed workflow.
+This article proposes a novel open-source, reusable, generalizable method\endnote{Our open-source tool is hosted at [REDACTED] for others to freely reuse.} to solve this problem.
+In short, it collects free open data plus a small amount of free data from the Google Maps API to train a local model to predict minimally-congested driving travel times with sparse data and simple computing hardware.
+\autoref{fig:workflow} shows a detailed workflow.
 
-\begin{figure*}[hbt!]
+\begin{figure*}[bt!]
     \centering
@@
 
-\begin{table*}[hbt!]
+\begin{table*}[tb!]
     \centering
-    \caption{Traffic control elements in the network model, as identified by OSM tags.}\label{tab:node_count}
+    \caption{Traffic control elements (as identified by OpenStreetMap tags) present in the graph, demonstrating their sparseness relative to the total intersections and nodes in the graph.}\label{tab:traffic_control_counts}
     \begin{tabular}{lr}
         \toprule
-        Traffic control element    & Node count \\ \midrule
-        Crossing                   &     21,559 \\
+        Element                        &   Count \\
+        \midrule
+        Crossing                       &  21,560 \\
         Stop sign                      &  26,304 \\
-        Turning circle             &     17,886 \\
-        Traffic signal             &     15,261 \\
-        Motorway junction          &      1,436 \\
+        Traffic signal                 &  15,262 \\
         Mini roundabout                &      44 \\
         Give way                       &     189 \\
-        3-way+ street intersection &    126,982 \\
-        Nodes without any tag      &    699,912 \\
-        Total nodes                &    782,825 \\ \bottomrule
+        \midrule
+        Total traffic control elements &  63,359 \\
+        Total street intersections     & 127,093 \\
+        Total nodes                    & 782,825 \\
+        \bottomrule
     \end{tabular}
@@
 
-We define the study area as the convex hull around the intersection of the Los Angeles County boundary and the Los Angeles urban area boundary from the Global Human Settlement Layer's Urban Center Database \citep{florczyk2019description, GHS2019}.
+We define an implementation study area as the convex hull around the intersection of the Los Angeles County boundary and the Los Angeles urban area boundary from the Global Human Settlement Layer's Urban Center Database \citep{florczyk2019description, GHS2019}.
 This allows us to retain the main urbanized area without adjacent metropolitan areas (such as the Inland Empire).
-We then model the drivable street network within this study area from OpenStreetMap using the OSMnx software \citep{boeing_modeling_2025}, retaining the strongly connected component, to generate a graph with 782,825 nodes.
-This graph is unsimplified, meaning that nodes represent intersections and dead-ends, as well as curving street line geometry vertices.
-Table~\ref{tab:node_count} summarizes these nodes' tagged traffic control elements: these data are sparse and most intersections lack traffic control information.
+We then model the drivable street network within this study area from OpenStreetMap using the OSMnx package \citep{boeing_modeling_2025} and retaining the strongly connected component, to produce an unsimplified graph with 782,825 nodes, of which 127,093 are street intersections.
+The graph also contains 63,359 tagged traffic control elements.
+\autoref{tab:traffic_control_counts} summarizes these nodes' tagged traffic controls: these data are sparse and many true intersections lack traffic control information.
 
 To generate origin-destination (OD) pairs for training a prediction model, we first over-sample 5,000,000 random node pairs from the street intersections and dead-ends in this graph.
-We then filter these down to realistic, minimally-congested trip patterns using the most-recently released Uber movement data \citep{ubermovement2020}.
-These data derive from from Uber trip GPS traces aggregated by tract-to-tract flow and hour of the day.
-We filter our OD pairs down to those that have a matching real-world trip ($n$ = 1,197,513) that occurred during the 03:00 hour ($n$ = 41,360) to best approximate minimally-congested traffic conditions.
+We then filter these down to realistic, minimally-congested trip patterns by using the most-recently released Uber movement data \citep{ubermovement2020}.
+These data derive from Uber trip GPS traces aggregated by tract-to-tract flow and hour of the day.
+We filter our OD pairs down to those that have a matching real-world trip ($n$ = 1,197,513) that occurred during the 03:00 hour ($n$ = 41,378) to best approximate minimally-congested traffic conditions.
+These Uber data are not a perfect measure of all real-world trips, but do validate the presence of travel demand between different OD pairs at different times of day.
 
-Finally, we collect \enquote{gold standard} travel times from the Google Maps API (specifically, the Routes API), which has been used in many studies to ground-truth travel times  \citep[e.g.,][]{ludwig2023traffic, hu2020estimating, wang2011estimating, fu2023comparative, delmelle2019travel}.
+Then we proxy \enquote{true travel times} by collecting travel times from the Google Maps Routes API, which has been used in numerous studies as a \enquote{gold standard} to proxy real-world travel times as discussed in the background section \citep[e.g.,][]{ludwig2023traffic, hu2020estimating, wang2011estimating, fu2023comparative, delmelle2019travel}.
 For each OD pair, the API provides the fastest network path, its travel time, and its length.
@@
 We used Google's \enquote{BEST\_GUESS} traffic-aware model to predict its \enquote{duration\_in\_traffic} travel time.
-The API was unable to solve 16 OD pairs' routes, resulting in 41,360 OD pairs with a Google travel time, which we used for prediction modeling.
+The API was unable to solve 18 OD pairs' routes, resulting in 41,360 OD pairs with a true travel time, which we use as the response in our prediction model.
 
-\begin{figure*}[hbt!]
+\begin{figure*}[tb!]
     \centering
     \includegraphics[width=1.0\textwidth]{fig_turns_definition.jpg}
-    \caption{Angular definition of turns in the model.}\label{fig:turns_definition}
+    \caption{Angular definition of turns in the prediction model.}\label{fig:turns_definition}
 \end{figure*}
 
-\subsection{Travel Time Prediction}
+\subsection{Model Specification and Tuning}
 
-Our travel time prediction model has two steps.
-First, we calculate \enquote{naïve travel times} across a set of trips using open data and open-source software.
-Second, we train a random forest regression model to predict the Google \enquote{gold standard} travel times from our \enquote{naïve travel times} plus a set of covariates.
-This subsection details this process.
+Conceptually, our travel time prediction model has two steps.
+First, we calculate \enquote{naïve} travel times across a set of trips using open data and open-source software.
+Second, we train a random forest (regression) model to predict our true travel times from our naïve travel times plus a set of covariates.
 
-To calculate naïve travel times, we solve each OD pair's shortest path using Dijkstra's algorithm to minimize edge traversal time at the speed limit.
-Then we count the number of traffic control elements (including traffic signals, pedestrian crossings, give ways, and mini roundabouts) and turns (i.e., left, slight left, right, slight right, and u-turn), as shown in Figure~\ref{fig:turns_definition}, along each OD pair's route.
+To calculate the naïve travel times for step one, we solve each OD pair's shortest path using Dijkstra's algorithm to minimize edge traversal time at the speed limit.
+Then we count the number of traffic control elements across the 5 types in \autoref{tab:traffic_control_counts} (stop signs, traffic signals, pedestrian crossings, give ways, and mini roundabouts), and the number of turns across the 5 types in \autoref{fig:turns_definition} (left, slight left, right, slight right, and u-turn), that were encountered along each OD pair's naïve route.
+Next we specify a travel time prediction model to predict an OD pair's true travel time (proxied by Google travel time) as a function of its naïve travel time plus these counts of each type of traffic control element.
+This model's specification is generalized by \autoref{eq:prediction_model}:
 
-Then we train a random forest regression model to predict a route's Google travel time from  the naïve travel time plus the count of each type of traffic control element and turn encountered along the naïve shortest path.
-We also tested other algorithms, including ordinary least squares, gradient boosting regression, AdaBoost regression, and decision tree regression, but settled on random forest regression as its ensemble accuracy substantially outperforms the others.
+\begin{equation}
+\label{eq:prediction_model}
+y = f(X) + \epsilon
+\end{equation}
 
-We divide our OD pairs into a standard 80/20 training/test split, then tune the hyperparameters through a randomized grid search with 5-fold cross-validation by calculating their mean absolute error (MAE), defined in Equation~\ref{eq:mae}:
+where $f$ is a prediction model to be trained on our data, $y$ is a length-$n$ response vector (true travel time), $X$ is a matrix of $n$ observations on 11 predictors (naïve travel time plus counts of the 5 types of traffic control element encountered and counts of the 5 types of turn directions encountered), and $\epsilon$ is random error.
+
+We train candidate models of this specification using several algorithms: decision trees, random forest, gradient boosting, and AdaBoost.
+A set of hyperparameters controls this algorithmic training.
+To tune each's hyperparameters for optimal performance, we split our data into a standard 80/20 training/test split then conduct a randomized grid search with 5-fold cross-validation to minimize the mean absolute error (MAE), defined by \autoref{eq:mae}:
 
@@
 \label{eq:mae}
-\text{MAE} = \frac{\sum^{n}_{i=1} \left|{t_i - \hat{t_i}}\right|}{n}
+\text{MAE} = \frac{\sum^{n}_{i=1} \left|{y_i - \hat{y_i}}\right|}{n}
 \end{equation}
 
-where $n$ is the total number of trips, $t_i$ is trip $i$'s observed travel time, and $\hat{t_i}$ is trip $i$'s predicted travel time.
+where $n$ is the total trip count, $y_i$ is trip $i$'s true travel time, and $\hat{y_i}$ is $i$'s predicted travel time.
 
-\subsection{Validation}
+\subsection{Model Selection and Validation}
 
-We validate our out-of-sample predictions against the \enquote{gold standard} Google travel times using five measures.
-First, we calculate the MAPE, following Equation~\ref{eq:mape}.
-Second, we calculate the MAE, following Equation~\ref{eq:mae}, to measure the average absolute difference in seconds between our prediction and the reference Google value.
-Third, we perform difference-in-means ($\delta$) $t$-tests to determine whether our predicted travel times are statistically significantly different (at the $p<=0.05$ significance level) from the reference Google values.
-Despite the large sample size, we expect insignificant $t$-statistics if the predicted travel times' distribution approximates the Google travel times' distribution.
-Fourth, we calculate the standard $R^2$ to measure our model's success in explaining travel time variation.
-Fifth and finally, we calculate the average of the pairwise ratios (APR) of our predicted travel time and the Google travel time to measure the average over- or under-prediction of the trips' travel times, as defined in Equation~\ref{eq:apr}:
+Once we have our candidate models tuned and trained, we select our final prediction model and validate the out-of-sample predictions against true travel times via six accuracy indicators.
+First, we calculate the MAPE, following \autoref{eq:mape}.
+Second, we calculate the MAE, following \autoref{eq:mae}.
+Third, we calculate the mean squared error (MSE), which is more sensitive to outliers than MAE or MAPE, following \autoref{eq:mse} (with the same variables as defined for \autoref{eq:mae}):
+
+\begin{equation}
+\label{eq:mse}
+\text{MSE} = \frac{\sum^{n}_{i=1} (y_i - \hat{y_i})^2}{n}
+\end{equation}
+
+Fourth, we perform difference-in-means ($\delta$) $t$-tests to determine whether our predicted travel times are statistically significantly different (at the 95\% confidence level) from the true travel times, to detect absolute systematic prediction bias.
+Despite the large sample size, we expect insignificant $t$-statistics if the predicted travel times' distribution closely matches the true travel times' distribution.
+Fifth, we calculate the average pairwise ratio (APR) of our predicted travel times and true travel times, to detect relative systematic prediction bias, as defined by \autoref{eq:apr} (with the same variables as defined for \autoref{eq:mae}):
 
@@
 \label{eq:apr}
-\text{APR} = \frac{\sum^{n}_{i=1}\frac{t_i}{\hat{t_i}}}{n}
+\text{APR} = \frac{\sum^{n}_{i=1}\frac{y_i}{\hat{y_i}}}{n}
 \end{equation}
 
-where $n$ is the total trip count, $t_i$ is trip $i$'s Google travel time, and $\hat{t_i}$ is our predicted travel time for trip $i$.
+Sixth and finally, we calculate the coefficient of determination, $R^2$, to measure our model's ability to explain observed travel time variation.
 
@@
 
-\subsection{Optimized Hyperparameters}
+\subsection{Optimized Hyperparameterization}
 
-Our final optimized model hyperparameterization uses 600 decision trees, random sampling with replacement in each decision tree, a maximum decision tree depth of 10, a non-limited maximum number of input explanatory features, a requirement of at least two samples to split at a decision node and one sample at a leaf node, and equal sample weighting.
-Finally, we check our trained model for overfitting by conducting 5-fold cross-validation across the whole sample using tuned hyperparameters: the resulting five MAE values (75.3, 74.5, 73.5, 73.3, 74.7) are all similar to each other, which indicates no overfitting problem as the hyperparameters are not overly specific to the training set.
+After tuning, the four model training algorithms perform similarly across most---but not all---of the accuracy indicators (\autoref{tab:validation_results}).
+They fall roughly in a similar range for MAPE (7.8--9.0\%), MAE (71.8--79.7 seconds), MSE (12,140--13,517 square seconds), APR (0.99--1.01), and $R^2$ (0.93), demonstrating the robustness of our model to algorithmic particulars.
+However, the random forest and decision tree models exhibit much lower absolute systematic prediction bias ($\delta$ = 0.45 and 0.44 seconds, respectively) than the gradient boosting and AdaBoost models do (-19.3 and -9.7 seconds, respectively).
+In other words, all these models have comparable mean errors---but the average random forest and decision tree predictions are statistically insignificantly different from the average true travel time, whereas gradient boosting and AdaBoost significantly under-predict it.
+
+Therefore, we select the random forest model as our \enquote{final} travel time prediction model because it predicts better overall (than the decision tree model) and better avoids systematic under-prediction (than the gradient boosting and AdaBoost models).
+This final model trains an ensemble of decision trees to predict unobserved response values by essentially dividing the data into boxes then making predictions based on the means of those boxes.
+Its optimized hyperparameterization uses 400 decision trees, random sampling with replacement in each decision tree, a maximum decision tree depth of 10, use of all available input explanatory features at each split, a requirement of at least two samples to split at a decision node, and the default settings that require at least one sample at a leaf node and apply equal sample weighting.
+
+Finally, we check this final model for overfitting by conducting another 5-fold cross-validation across the whole sample using the tuned hyperparameters.
+The resulting five MAE values (75.3, 74.5, 73.5, 73.3, 74.6) are all very close to each other, indicating that the model is not overfitted and its hyperparameters are not overly specific to the training set.
 
@@
 
-Table~\ref{tab:validation_results} details our travel time prediction model's out-of-sample improvement over the initial naïve travel time calculation.
-Our model achieves a MAPE of 8.4\%, in line with the approximately 10\% MAPEs seen in the state-of-the-art travel time prediction literature, but without their extensive and expensive input data requirements.
-In comparison, our initial naïve travel time calculation yields a much worse MAPE of 21.1\%.
-Whereas the naïve travel time calculation's MAE is 183.5 seconds from the Google travel time, our predicted travel times' MAE is just 75.2 seconds---an improvement by a factor of 2.4.
-
 \begin{table*}[tb!]
-    \centering
-    \caption{Results of out-of-sample (testing set) prediction of (1) our prediction model and (2) the initial naïve travel time calculation, validated against the corresponding Google travel times.
-The five validation measures are (1) MAPE, (2) MAE, (3) difference-in-means ($\delta$) and corresponding $t$-test $p$-value, (4) APR, and (5) $R^2$.
+    \centering\small
+    \caption{Out-of-sample prediction accuracy of our chosen model (random forest, bold) versus the initial naïve travel time model and the discarded alternative models, all validated against the corresponding true travel times.
+$n$ = 8,272 for each.
+The six accuracy indicators are (1) MAPE \%, (2) MAE in seconds, (3) MSE in square seconds, (4) difference-in-means ($\delta$) in seconds and its corresponding $t$-test's $p$-value, (5) APR, and (6) $R^2$.
 See methods for definitions.}\label{tab:validation_results}
-    \begin{tabular}{lrr}
+    \begin{tabular}{lrrrrrrr}
         \toprule
-                                      & Prediction model & Naïve calculation \\
+        Model               & MAPE  & MAE    & MSE      & $\delta$ & $p$  & APR  & $R^2$ \\
         \midrule
-        MAPE (\%)                     &             8.43 &             21.13 \\
-        MAE (seconds)                 &            75.23 &            183.55 \\
-        $\delta$ (seconds)            &             0.38 &           -182.71 \\
-        $p$-value                     &             0.75 &             <0.01 \\
-        APR                           &             1.01 &              0.79 \\
-        $R^2$                         &             0.93 &              0.74 \\
-        \midrule
-        $n$                           &           8,272 &            8,272 \\
+        Initial naïve       & 21.13 & 183.55 & 48155.43 &  -182.71 & <0.01 & 0.79 & 0.74  \\
+        \textbf{Random forest}       &  \textbf{8.43} &  \textbf{75.24} & \textbf{12140.14} &     \textbf{0.45} &  \textbf{0.71} & \textbf{1.01} & \textbf{0.93}  \\
+        Gradient boosting   &  7.84 &  71.81 & 12572.83 &   -19.30 & <0.01 & 0.99 & 0.93  \\
+        Decision trees      &  8.96 &  79.74 & 13517.03 &     0.44 &  0.73 & 1.01 & 0.93  \\
+        AdaBoost            &  8.21 &  74.05 & 12416.08 &    -9.73 & <0.01 & 1.00 & 0.93  \\
         \bottomrule
@@
 
-The $t$-test reveals significant differences between the initial naïve travel time calculations and the Google travel times: the $\delta$ of -182.7 seconds corresponds to $p<0.01$.
-In other words, the naïve travel time under-predicts Google by over 3 minutes.
-However, the $t$-test reveals insignificant differences between our model's out-of-sample predictions and the Google travel times: the $\delta$ of 0.4 seconds corresponds to a $p$ value of 0.75.
+Compared to the initial naïve travel time calculation, our travel time prediction model exhibits substantial out-of-sample improvement (\autoref{tab:validation_results}).
+Our model produces a MAPE of 8.4\%, in line with the \textasciitilde10\% MAPEs seen in the state-of-the-art travel time prediction literature, but without their extensive and expensive input data requirements.
+In comparison, our initial naïve travel time calculation produces a much worse MAPE of 21.1\%.
+Whereas the naïve travel time's MAE is 183.6 seconds, our model's predicted travel times' MAE is just 75.2 seconds---an improvement by a factor of 2.4.
+Similarly for MSE, our model improved by a factor of 4.0.
+
+The $t$-test reveals significant differences between the initial naïve travel time calculations and the true travel times: the $\delta$ of -183.6 seconds corresponds to $p$ < 0.01.
+In other words, the naïve travel time under-predicts true travel time by over 3 minutes.
+However, the $t$-test reveals insignificant differences between our model's out-of-sample predictions and the true travel times: the $\delta$ of 0.45 seconds corresponds to a $p$ value of 0.71.
 That is, our prediction model over-predicts by less than half a second on average, which is statistically insignificantly different from zero.
-If we compare the aforementioned MAEs to these differences-in-means, we can see that the initial naïve travel time calculations significantly and consistently under-predict travel time, but our prediction model shows no such directional bias: its random error balances out between (smaller absolute) over-and under-prediction.
+If we reconsider the aforementioned MAEs in light of these differences-in-means, we can see that the initial naïve travel time calculations significantly and consistently under-predict travel time, but our prediction model shows no such directional bias: its random error averages out between (smaller absolute) over-and under-prediction.
 
 The APR offers another lens on this finding.
-On average across the OD pairs, the initial naïve travel time calculation under-predicts Google travel time by 21\%, but our prediction model over-predicts it by just 1\%.
+On average across the OD pairs, the initial naïve travel time calculation under-predicts true travel time by 21\%, but our prediction model over-predicts it by just 1\%.
 Our model also explains more of the variation in travel time: its $R^2$ of 0.93 is substantially higher than the 0.74 $R^2$ of the initial naïve travel time calculation.
-In summary, each of our five validation measures demonstrates that our prediction model drastically improves on naïve travel time calculations---without needing extensive, expensive input data or advanced deep learning software---and offers travel time predictions much closer to those of Google's \enquote{gold standard} predictions.
+In sum, each of our six accuracy indicators demonstrates that our prediction model drastically improves on naïve travel time calculations---without needing extensive, expensive input data or advanced deep learning software---and offers much more accurate travel time predictions.
 
@@
 
-Re-state the problem.
-This study set out to investigate if a middle-ground travel time prediction model was possible.
+The planning literature shows that planners often use driving travel time prediction techniques far below the state-of-the-art due to cost, data, and technical capacity constraints.
+This study investigated if a better middle-ground travel time prediction method was feasible.
 It identified three criteria for success: (1) it should use free software and data and avoid bespoke sensor data collection; (2) it should be easy and accessible to use; (3) it should offer better accuracy than naïve predictions (e.g., minimizing Euclidean distance, network distance, street edge traversal time) and better spatial resolution than zone-to-zone travel times.
@@
 Rather, it empowers the scholars and practitioners who need it the most---those working on the frontlines of urban planning and policymaking.
-These scholars and practitioners rely on simple but inaccurate routing models---such as minimizing Euclidean distance, network distance, or edge traversal time---when they lack the technical expertise, computing equipment, or extensive data needed to implement the literature's cutting-edge models.
-However, this study demonstrated how these naïve methods systematically under-predict real-world travel times.
-Our model addresses this problem by incorporating sparse open data on traffic control elements and turns into the prediction and achieves an accuracy in line with state-of-the-art methods---but without their data needs or computing requirements.
-In other words, our model does not replace the state-of-the-art methods for those with extensive resources available to implement them, but rather offers far more accurate predictions for urban planners who would otherwise fall back on simpler naïve models.
+The literature shows that these scholars and practitioners often rely on simple but inaccurate routing models---such as minimizing Euclidean distance, network distance, or edge traversal time---when they lack the technical expertise, computing equipment, or extensive data needed to implement the literature's cutting-edge models.
+However, this study demonstrates how these naïve methods systematically under-predict real-world travel times.
+Our model addresses this problem by incorporating sparse open data on traffic control elements and turns into the prediction to achieve an accuracy in line with state-of-the-art methods---but without their data needs or computing requirements.
+In other words, our model does not replace the state-of-the-art congested-traffic methods for those with extensive resources available to implement them, but rather offers far more accurate predictions for urban planners who would otherwise fall back on simpler naïve models.
 Inaccurate travel times skew empirical understandings and mislead planning interventions.
-Our model represents an important contribution for urban planning scholarship and evidence-based practitioner interventions.
+Our model represents an important middle-ground contribution for urban planning scholarship and evidence-informed practitioner interventions.
 
@@
 Our implementation focused on Los Angeles but is not inherently tied to it, and future work should emphasize its generalizability in cities with potentially sparser or lower-quality open data.
-We trained this model as a proof-of-concept for minimally-congested times of day, but the same concept can be extended with similar training data for any other time of day.
+We trained this model as a proof-of-concept for minimally-congested times of day, but the same concept can be extended with similar training data for any other time of day if congestion data exist.
 Future research should also expand the kinds of training and validation data.
@@
 Travel time prediction is central to questions of urban accessibility, mode and route choice, and individuals' location decisions.
-True congested travel time vary drastically throughout the day and require large volumes of real-time or proprietary data, as well as complex algorithms, to model.
+True congested travel times vary drastically throughout the day and require large volumes of real-time or proprietary data, as well as complex algorithms, to model.
 These technical challenges and costs present a hurdle for many urban planners, who often turn to simpler models with lower and often free data and computing requirements.
-The traditional naïve methods are easy to implement, but produce inaccurate predictions.
+These traditional naïve methods may be easy to implement, but they produce wildly inaccurate predictions.
 
-This article introduced a middle-ground method to simply but accurately predict minimally-congested travel time using free data.
+This article introduced a better middle-ground method to simply but accurately predict minimally-congested driving travel time using free data.
 It improves on traditional, common, naïve predictions by training on a small one-off collection of free but proprietary high-quality Google data on travel times.
-Also of note is the sparseness of our open data: OpenStreetMap contains little information on traffic controls, as only around 28\% of the intersections in the study area have traffic control information.
-Even so, our prediction model using random forest regression still predicts Google travel times with high accuracy.
+Also of note is the sparseness of our open data: OpenStreetMap contains only sporadic information on traffic controls' presence.
+Even so, our prediction model demonstrates very high accuracy.
 This offers planning practitioners and scholars a better method to predict travel times for free with much better accuracy than traditional naïve methods offered.
@@
 
-The authors wish to thank Youngseo Kweon and Jaehyun Ha for additional research assistance.
+[REDACTED]
+%The authors wish to thank Youngseo Kweon and Jaehyun Ha for additional research assistance.
 

### 2025-05-03 tool/01-preprocess_input_data.py
Handling multiple tags for one node scenerio
Refactored the code to handle multiple tags for one node scenerio and fixed minor bugs

@@ -72,7 +72,7 @@ def get_annotated_graph(
     ca_tract_gdf: gpd.GeoDataFrame,
     uber_tract_gdf: gpd.GeoDataFrame,
 ) -> gpd.GeoDataFrame:
-    """Annotate information to all nodes in the given .
+    """Annotate information to all nodes in the given.
 
     Parameters
     ----------
@@ -103,13 +103,18 @@ def get_annotated_graph(
     ]
 
     convex_polygon = gpd.GeoDataFrame(geometry=polygon).iloc[0]["geometry"]
-    graph = ox.graph_from_polygon(convex_polygon, simplify=False, network_type="drive")
+    graph_sim = ox.graph_from_polygon(convex_polygon, simplify=True, network_type="drive")
+    graph_nonsim = ox.graph_from_polygon(convex_polygon, simplify=False, network_type="drive")
 
     # Get strongly connected graph
-    connected_graph = ox.truncate.largest_component(graph, strongly=True)
-    ox.save_graphml(connected_graph, filepath=constants.LA_CLIP_CONVEX_NETWORK_GML_FILE_PATH)
+    connected_graph_sim = ox.truncate.largest_component(graph_sim, strongly=True)
+    connected_graph_nonsim = ox.truncate.largest_component(graph_nonsim, strongly=True)
+    ox.save_graphml(
+        connected_graph_nonsim,
+        filepath=constants.LA_CLIP_CONVEX_NETWORK_GML_FILE_PATH,
+    )
 
-    gdf_nodes, gdf_edges = ox.graph_to_gdfs(connected_graph)
+    gdf_nodes, gdf_edges = ox.graph_to_gdfs(connected_graph_sim)
     gdf = gdf_nodes.reset_index(drop=False)
     gdf_proj = ox.projection.project_gdf(gdf, to_latlong=True)
     gdf_proj["x"] = gdf_proj["geometry"].x

### 2025-05-05 paper/main.tex
add data availability statement
add conflict of interest statement

@@
 % !TeX program = pdflatex
-% Minimally-Congested Travel Time Prediction from Sparse Open Data
+% Travel Time Prediction from Sparse Open Data
 % Authors: Geoff Boeing and Yuquan Zhou
@@
 
-\newcommand{\myname}{Geoff Boeing and Yuquan Zhou}
+\newcommand{\myname}{Redacted for review}%Geoff Boeing and Yuquan Zhou}
 \newcommand{\myemail}{boeing@usc.edu}
-\newcommand{\myaffiliation}{University of Southern California}
+\newcommand{\myaffiliation}{Redacted for review}%University of Southern California}
 \newcommand{\paperdate}{2025}
-\newcommand{\papertitle}{Minimally-Congested Travel Time Prediction from Sparse Open Data}
+\newcommand{\papertitle}{Travel Time Prediction from Sparse Open Data}
 \newcommand{\papercitation}{Boeing, G. and Y. Zhou. \paperdate. \papertitle. Under review at \emph{Journal Name}.}
@@
 \title{\papertitle}
-\author[]{Redacted for review}%{\myname}
-\affil[]{Redacted for review}%{\myaffiliation}
+\author[]{\myname}
+\affil[]{\myaffiliation}
 \date{}%\paperdate}
@@
 
-Travel time prediction is central to accessibility analyses, sustainable transportation infrastructure provision, and active transportation interventions.
-However, calculating accurate travel times, especially for driving, requires either extensive technical capacity and bespoke data, or sources like the Google Maps API that quickly become prohibitively expensive to analyze thousands or millions of trips necessary for metropolitan-scale planning.
-These obstacles make it difficult for researchers, practitioners, and community advocates to develop an evidence base for more sustainable transportation and equitable accessibility.
-This article introduces a free, open-source minimally-congested driving time prediction model with minimal cost, data, and computational requirements.
+Travel time prediction is central to transport geography and planning's accessibility analyses, sustainable transportation infrastructure provision, and active transportation interventions.
+However, calculating accurate travel times, especially for driving, requires either extensive technical capacity and bespoke data, or sources like the Google Maps API that quickly become prohibitively expensive to analyze thousands or millions of trips necessary for metropolitan-scale analyses.
+Such obstacles particularly challenge less-resourced researchers, practitioners, and community advocates.
+This article argues that a middle-ground is needed to provide reasonably accurate travel time predictions without extensive data or computing requirements.
+It introduces a free, open-source minimally-congested driving time prediction model with minimal cost, data, and computational requirements.
 It trains and tests this model using the Los Angeles, California urban area as a case study by calculating naïve travel times from open data then developing a random forest model to predict true travel times as a function of those naïve times plus open data on turns and traffic controls.
-Validation shows that this method produces accurate, free, and simple travel time predictions, offering a superior middle-ground technique for urban planners.
+Validation shows that this method offers a superior middle-ground technique that balances reasonable accuracy with minimal resource requirements.
 
@@
 
-Travel time prediction underpins many transport planning processes and research problems.
+Travel time prediction underpins many transport planning processes and geographical research problems.
 It is key to understanding urban accessibility, transport mode and route choice, and location decisions \citep{zhang2015gradient,jenelius2013travel}.
-Predicting congested travel times (i.e., when traffic congestion impedes flows) requires large volumes of real-time data on traffic conditions \citep[e.g.,][]{hou2018network}.
+Predicting congested travel times (i.e., when traffic congestion impedes flows) requires large volumes of real-time geospatial data on traffic conditions \citep[e.g.,][]{hou2018network}.
 Such data are proprietary, expensive, and beyond the reach of many planning practitioners and scholars \citep{carrion2012value,wang2011estimating}.
-This is usually not a problem for well-resourced organizations with highly-skilled labor---however, technical expertise, complex data and software requirements, and high costs pose insurmountable obstacles for many urban planning researchers, practitioners, and community advocates \citep{giles2022creating}.
+This is usually not a problem for well-resourced organizations with highly-skilled labor---however, GIScience technical expertise, complex data and software requirements, and high costs pose insurmountable obstacles for many urban planning researchers, practitioners, and community advocates \citep{giles2022creating}.
 
-Although typical \enquote{naïve} travel time prediction methods (which minimize travel distance or street segment traversal time) provide easy solutions, they ignore delays from turns and traffic controls and thus usually under-predict real-world driving travel time \citep{ludwig2023traffic,salonen2013modelling,yiannakoulias2013estimating}.
-However, as we will argue, minimally-congested travel time prediction offers a better middle ground of more-accurate predictions than naïve travel times do---yet much lower data and computing requirements than real-time traffic-aware congested travel time prediction.
+Travel time is fundamentally a function of distance and speed, which itself is a function of speed limits, traffic controls, turns, and congestion.
+Although common \enquote{naïve} travel time prediction methods (which minimize travel distance or street segment traversal time) provide easy solutions, they ignore the other components and thus usually under-predict real-world driving travel time \citep{ludwig2023traffic,salonen2013modelling,yiannakoulias2013estimating}.
+However, as we will argue, minimally-congested travel time prediction offers an important yet under-realized middle ground: much more accurate predictions than naïve travel times, yet much lower data and computing requirements than real-time traffic-aware congested travel time prediction.
 
-To address this need for a middle ground, this article contributes a novel accurate, easy-to-use, and free method of minimally-congested driving travel time prediction.
+To address the need for such a middle ground, this article contributes a novel accurate, easy-to-use, and free method of minimally-congested driving travel time prediction.
 Our results show a substantial improvement over traditional, naïve methods of travel time prediction common in the urban planning literature.
-As we shall show in this article, across the Los Angeles urban area, our average trip's travel time prediction differs from Google's \enquote{gold standard} prediction by only 0.45 seconds.
-In contrast, simple---but common---naïve travel time prediction minimizing max-speed edge traversal times under-predicts by 183.55 seconds on average.
+As we shall show in this article, across the Los Angeles urban area, our average trip's travel time prediction differs from Google's \enquote{gold standard} prediction by only 0.38 seconds.
+In contrast, simple---but common in the literature---naïve travel time prediction minimizing max-speed edge traversal times under-predicts by 183.69 seconds on average.
 
@@
 First it reviews the state of the art in travel time prediction and then considers the less-sophisticated methods that are standard in the current planning literature.
-Next it describes our data sources, models, and validation techniques.
+Next it describes our proposed model, data sources, and validation techniques.
 Then it presents our findings, highlighting both the feasibility and quality of our predictions in relation to simple naïve predictions or difficult expensive predictions.
@@
 Modern travel time prediction techniques tend to fall into one of two camps: (1) sophisticated, accurate, but difficult to execute; or (2) naïve, less-accurate, but simple to execute.
-The former is the state-of-the-art and has been the subject of much work in the recent computer science and engineering literatures.
+The former is the state-of-the-art and has been the subject of much GIScience work particularly in the recent computer science and engineering literatures.
 The latter tends to appear often in urban planning research and practice.
@@
 
-A better way would be to seek a middle ground between expensive, complicated, point-to-point empirical travel times and overly-simple naïve predictions.
-Such free empirical travel time predictions may not reflect congested travel times (which vary drastically throughout the day and week), but would provide a good prediction of minimally-congested travel times.
+Travel time is essentially a function of distance, speed limits, traffic controls, turns, and traffic congestion.
+Naïve methods tend to predict it from just the first one or two of these components, whereas resource-intensive state-of-the-art methods predict it from all five.
+But, with open data, the first four components can now be freely acquired in many parts of the world---only congestion data remains universally proprietary, expensive, and inaccessible.
+Therefore, could there be a middle-ground between expensive, complicated, real-time predictions and overly-simple naïve predictions?
+Can we predict travel times from free open data on distance, speed limits, traffic controls, and turns?
+Such predictions may not reflect real-time congestion (which varies drastically throughout the day and week), but would accurately represent minimally-congested travel times.
+
 This middle-ground method should meet three criteria.
@@
 Second, it should be a simple and accessible method for urban planners without advanced technical skills.
-Third, it should provide a substantial improvement in accuracy or spatial resolution over naïve predictions such as minimizing Euclidean distance, network distance, street edge traversal time, and exsiting TAZ-to-TAZ travel times, and approach the accuracy of the state-of-art models.
-Meeting these three criteria, a middle ground should be inexpensive, easy-to-use, and reasonably accurate.
+Third, it should provide a substantial improvement in accuracy and spatial resolution over naïve predictions such as minimizing Euclidean distance, network distance, street edge traversal time, and exsiting TAZ-to-TAZ travel times, and approach the accuracy of the state-of-art models.
+Meeting these three criteria, a middle ground would be inexpensive, easy-to-use, and reasonably accurate.
 
@@
 
-This article proposes a novel open-source, reusable, generalizable method\endnote{Our open-source tool is hosted at [REDACTED] for others to freely reuse.} to solve this problem.
+This article proposes a novel open-source, reusable, generalizable method\endnote{Our open-source code is hosted at [REDACTED] for others to freely reuse.} to solve this problem.
 In short, it collects free open data plus a small amount of free data from the Google Maps API to train a local model to predict minimally-congested driving travel times with sparse data and simple computing hardware.
@@
 After tuning, the four model training algorithms perform similarly across most---but not all---of the accuracy indicators (\autoref{tab:validation_results}).
-They fall roughly in a similar range for MAPE (7.8--9.0\%), MAE (71.8--79.7 seconds), MSE (12,140--13,517 square seconds), APR (0.99--1.01), and $R^2$ (0.93), demonstrating the robustness of our model to algorithmic particulars.
-However, the random forest and decision tree models exhibit much lower absolute systematic prediction bias ($\delta$ = 0.45 and 0.44 seconds, respectively) than the gradient boosting and AdaBoost models do (-19.3 and -9.7 seconds, respectively).
+They fall roughly in a similar range for MAPE (7.9--9.0\%), MAE (72.0--80.0 seconds), MSE (12,155--13,570 square seconds), APR (0.99--1.01), and $R^2$ (0.93), demonstrating the robustness of our model to algorithmic particulars.
+However, the random forest and decision tree models exhibit much lower absolute systematic prediction bias ($\delta$ = 0.38 and 0.13 seconds, respectively) than the gradient boosting and AdaBoost models do (-19.2 and -9.8 seconds, respectively).
 In other words, all these models have comparable mean errors---but the average random forest and decision tree predictions are statistically insignificantly different from the average true travel time, whereas gradient boosting and AdaBoost significantly under-predict it.
+Therefore, we select the random forest model as our \enquote{final} travel time prediction model because it predicts better overall (than the decision tree model) and better avoids systematic under-prediction (than the gradient boosting and AdaBoost models).
 
-Therefore, we select the random forest model as our \enquote{final} travel time prediction model because it predicts better overall (than the decision tree model) and better avoids systematic under-prediction (than the gradient boosting and AdaBoost models).
 This final model trains an ensemble of decision trees to predict unobserved response values by essentially dividing the data into boxes then making predictions based on the means of those boxes.
 Its optimized hyperparameterization uses 400 decision trees, random sampling with replacement in each decision tree, a maximum decision tree depth of 10, use of all available input explanatory features at each split, a requirement of at least two samples to split at a decision node, and the default settings that require at least one sample at a leaf node and apply equal sample weighting.
-
 Finally, we check this final model for overfitting by conducting another 5-fold cross-validation across the whole sample using the tuned hyperparameters.
-The resulting five MAE values (75.3, 74.5, 73.5, 73.3, 74.6) are all very close to each other, indicating that the model is not overfitted and its hyperparameters are not overly specific to the training set.
+The resulting five MAE values (75.3, 74.5, 73.6, 73.2, 74.7) are all very similar, indicating that the model is not overfitted and its hyperparameters are not overly specific to the training set.
 
@@
 See methods for definitions.}\label{tab:validation_results}
+    \tabularnums{\small
     \begin{tabular}{lrrrrrrr}
@@
         \midrule
-        Initial naïve       & 21.13 & 183.55 & 48155.43 &  -182.71 & <0.01 & 0.79 & 0.74  \\
-        \textbf{Random forest}       &  \textbf{8.43} &  \textbf{75.24} & \textbf{12140.14} &     \textbf{0.45} &  \textbf{0.71} & \textbf{1.01} & \textbf{0.93}  \\
-        Gradient boosting   &  7.84 &  71.81 & 12572.83 &   -19.30 & <0.01 & 0.99 & 0.93  \\
-        Decision trees      &  8.96 &  79.74 & 13517.03 &     0.44 &  0.73 & 1.01 & 0.93  \\
-        AdaBoost            &  8.21 &  74.05 & 12416.08 &    -9.73 & <0.01 & 1.00 & 0.93  \\
+        Initial naïve       & 21.15 & 183.69 & 48214.06 &  -182.85 & <0.01 & 0.79 & 0.74  \\
+        \textbf{Random forest}       &  \textbf{8.42} &  \textbf{75.32} & \textbf{12154.79} &     \textbf{0.38} &  \textbf{0.76} & \textbf{1.01} & \textbf{0.93}  \\
+        Gradient boosting   &  7.86 &  71.99 & 12630.03 &   -19.20 & <0.01 & 0.99 & 0.93  \\
+        Decision trees      &  9.00 &  80.04 & 13570.17 &     0.13 &  0.92 & 1.01 & 0.93  \\
+        AdaBoost            &  8.20 &  74.04 & 12424.14 &    -9.76 & <0.01 & 1.00 & 0.93  \\
         \bottomrule
-    \end{tabular}
+    \end{tabular}}
 \end{table*}
@@
 Our model produces a MAPE of 8.4\%, in line with the \textasciitilde10\% MAPEs seen in the state-of-the-art travel time prediction literature, but without their extensive and expensive input data requirements.
-In comparison, our initial naïve travel time calculation produces a much worse MAPE of 21.1\%.
-Whereas the naïve travel time's MAE is 183.6 seconds, our model's predicted travel times' MAE is just 75.2 seconds---an improvement by a factor of 2.4.
+In comparison, our initial naïve travel time calculation produces a much worse MAPE of 21.2\%.
+Whereas the naïve travel time's MAE is 183.7 seconds, our model's predicted travel times' MAE is just 75.3 seconds---an improvement by a factor of 2.4.
 Similarly for MSE, our model improved by a factor of 4.0.
 
-The $t$-test reveals significant differences between the initial naïve travel time calculations and the true travel times: the $\delta$ of -183.6 seconds corresponds to $p$ < 0.01.
+The $t$-test reveals significant differences between the initial naïve travel time calculations and the true travel times: the $\delta$ of -182.9 seconds corresponds to $p$ < 0.01.
 In other words, the naïve travel time under-predicts true travel time by over 3 minutes.
-However, the $t$-test reveals insignificant differences between our model's out-of-sample predictions and the true travel times: the $\delta$ of 0.45 seconds corresponds to a $p$ value of 0.71.
+However, the $t$-test reveals insignificant differences between our model's out-of-sample predictions and the true travel times: the $\delta$ of 0.38 seconds corresponds to a $p$ value of 0.76.
 That is, our prediction model over-predicts by less than half a second on average, which is statistically insignificantly different from zero.
@@
 
+Travel time is fundamentally a function of distance, speed limits, traffic controls, turns, and congestion.
+Naïve methods predict only from the first one or two of these components.
+Resource-intensive state-of-the-art methods predict it from all five.
 The planning literature shows that planners often use driving travel time prediction techniques far below the state-of-the-art due to cost, data, and technical capacity constraints.
-This study investigated if a better middle-ground travel time prediction method was feasible.
-It identified three criteria for success: (1) it should use free software and data and avoid bespoke sensor data collection; (2) it should be easy and accessible to use; (3) it should offer better accuracy than naïve predictions (e.g., minimizing Euclidean distance, network distance, street edge traversal time) and better spatial resolution than zone-to-zone travel times.
+In this study, we propose that a better middle-ground travel time prediction method is feasible and identify three criteria for success: (1) it should use free software and data and avoid bespoke sensor data collection; (2) it should be easy and accessible to use; (3) it should offer better accuracy than naïve predictions (e.g., minimizing Euclidean distance, network distance, street edge traversal time) and better spatial resolution than zone-to-zone travel times.
 
@@
 This middle ground is not necessarily for computer scientists, high-end labs with top-of-the-line computing hardware, or well-resourced agencies with extensive data budgets.
-Rather, it empowers the scholars and practitioners who need it the most---those working on the frontlines of urban planning and policymaking.
+Rather, it empowers the scholars and practitioners who need GIScience solutions the most---those working on the frontlines of urban planning and policymaking.
 The literature shows that these scholars and practitioners often rely on simple but inaccurate routing models---such as minimizing Euclidean distance, network distance, or edge traversal time---when they lack the technical expertise, computing equipment, or extensive data needed to implement the literature's cutting-edge models.
@@
 We trained this model as a proof-of-concept for minimally-congested times of day, but the same concept can be extended with similar training data for any other time of day if congestion data exist.
+This is, of course, a big \enquote{if}---and the whole reason why minimally-congested predictions are valuable in the first place.
 Future research should also expand the kinds of training and validation data.
@@
 Travel time prediction is central to questions of urban accessibility, mode and route choice, and individuals' location decisions.
-True congested travel times vary drastically throughout the day and require large volumes of real-time or proprietary data, as well as complex algorithms, to model.
+True congested travel times vary drastically throughout the day and require large volumes of real-time or proprietary geospatial data, as well as complex algorithms, to model.
 These technical challenges and costs present a hurdle for many urban planners, who often turn to simpler models with lower and often free data and computing requirements.
@@
 This offers planning practitioners and scholars a better method to predict travel times for free with much better accuracy than traditional naïve methods offered.
+
+\section*{Data Availability Statement}
+
+The data that support the findings of this study are available at [OFFICIAL REPOSITORY URL REDACTED FOR ANONYMOUS PEER REVIEW.\@ REVIEWERS, PLEASE USE THIS ANONYMIZED URL FOR YOUR REVIEW INSTEAD:\@ https://figshare.com/s/0acd076b1a11286ff3e4].
+%https://github.com/gboeing/travel-time-prediction
+
+\section*{Conflict of Interest Statement}
+
+The authors have no relevant financial or non-financial competing interests to report.
 

### 2025-05-07 one_click_run.sh
Add shell script
With bug fix, and table 1 export

@@ -1,0 +1,14 @@
+#!/bin/zsh
+
+#conda env create -f environment.yml
+conda activate ttp
+
+if [ "$1" = "--requery" ]
+  then echo "requery the data..."
+  python3 tool/01-preprocess_input_data.py
+  python3 tool/02-routes_api.py "$2" "$3"
+fi
+
+echo "routing and modeling..."
+python3 tool/03-network_routing.py
+python3 tool/04-modeling.py

### 2025-05-07 tool/01-preprocess_input_data.py

@@ -21,7 +21,7 @@ def get_gdf_inputs() -> tuple[
 
     Returns
     -------
-        Tuple of GeoDataFrames with same CRS.
+        Tuple of GeoDataFrames with the same CRS.
 
     """
     ucdb_gdf = gpd.read_file(constants.UCDB_FILE_PATH)
@@ -221,15 +221,12 @@ def sample_od_pairs(
         .apply(lambda x: x.sample(1, random_state=123))
         .reset_index(drop=True)
     )
+    uber_dedup_merge_sample = uber_dedup_merge_sample[uber_dedup_merge_sample["hod"] == sample_hour]
 
-    sampled_hour_df = (
-        uber_dedup_merge_sample[uber_dedup_merge_sample["hod"] == sample_hour]
-        .sample(
-            n=len(uber_dedup_merge_sample) if sample_size is None else sample_size,
-            random_state=123,
-        )
-        .copy()
-    )
+    sampled_hour_df = uber_dedup_merge_sample.sample(
+        n=len(uber_dedup_merge_sample) if sample_size is None else sample_size,
+        random_state=123,
+    ).copy()
     sampled_hour_df.to_csv(constants.SAMPLED_OD_SAMPLE_HOUR_FILE_PATH)
     with Path.open(constants.SAMPLED_OD_SAMPLE_HOUR_PICKLE_PATH, "wb") as file:
         pickle.dump(sampled_hour_df, file)

### 2025-05-07 tool/03-network_routing.py

@@ -4,6 +4,7 @@
 
 import multiprocessing as mp
 import warnings
+from collections import Counter
 from heapq import heappop, heappush
 from itertools import count
 
@@ -603,6 +604,42 @@ def run() -> None:
     od_pair_sample = pd.read_csv(constants.SAMPLED_OD_ROUTES_API_FILE_PATH)
 
     graph = ox.io.load_graphml(constants.LA_CLIP_CONVEX_NETWORK_GML_FILE_PATH)
+    resim_graph = ox.simplification.simplify_graph(graph)
+    simplify_nodes, simplify_edges = ox.graph_to_gdfs(resim_graph)
+    # statistics for table 1
+    tag_list = ["traffic_signals", "stop", "crossing", "give_way", "mini_roundabout"]
+    tag_counter = Counter()
+
+    for _, data in graph.nodes(data=True):
+        highway = data.get("highway")
+        tags = []
+        if isinstance(highway, str):
+            tags = [t.strip() for t in highway.split(";")]
+            # if there are multiple tags in the node, split them into a list
+            # or if there is only one tag in the node, also put it as a list
+        elif isinstance(highway, list):
+            # if it is already a list, then pass it along
+            tags = highway
+        for tag in tags:
+            if tag in tag_list:
+                tag_counter[tag] += 1
+    total_controls = sum(tag_counter.values())
+    total_nodes = len(graph.nodes)
+    total_intersections = len(simplify_nodes[simplify_nodes["street_count"] > 1])
+
+    # Create table data
+    tc_rows = [
+        {"Element": "Crossing", "Count": tag_counter.get("crossing", 0)},
+        {"Element": "Stop sign", "Count": tag_counter.get("stop", 0)},
+        {"Element": "Traffic signal", "Count": tag_counter.get("traffic_signals", 0)},
+        {"Element": "Mini roundabout", "Count": tag_counter.get("mini_roundabout", 0)},
+        {"Element": "Give way", "Count": tag_counter.get("give_way", 0)},
+        {"Element": "Total traffic control elements", "Count": total_controls},
+        {"Element": "Total street intersections", "Count": total_intersections},
+        {"Element": "Total nodes", "Count": total_nodes},
+    ]
+    tc_df = pd.DataFrame(tc_rows)
+    tc_df.to_csv(constants.TRAFFIC_CONTROL_FILE_PATH, index=False)
 
     valid_nodes = set(graph.nodes)
 

### 2025-05-07 tool/constants.py

@@ -31,6 +31,8 @@ NETWORK_ROUTING_RESULT_FILE_PATH = (
     INTERMEDIATE_DIR + "OD3am_routes_api_network_routing_multiple_all.csv"
 )
 
+TRAFFIC_CONTROL_FILE_PATH = OUTPUT_DIR + "table1_traffic_controls.csv"
+
 RF_PREDICTION_RESULT_FILE_PATH = OUTPUT_DIR + "rf_prediction_result.csv"
 GB_PREDICTION_RESULT_FILE_PATH = OUTPUT_DIR + "gb_prediction_result.csv"
 AB_PREDICTION_RESULT_FILE_PATH = OUTPUT_DIR + "ab_prediction_result.csv"

### 2025-05-07 tool/04-modeling.py
add table 2 output

@@ -442,3 +442,21 @@ if __name__ == "__main__":
     adaboost(result)
     logger.info("adaboost done.")
     naive(result)
+
+    # combine evaluation result
+    file_paths = [
+        constants.NETWORK_ROUTING_EVALUATION_RESULT_FILE_PATH,
+        constants.RF_EVALUATION_RESULT_FILE_PATH,
+        constants.GB_EVALUATION_RESULT_FILE_PATH,
+        constants.DT_EVALUATION_RESULT_FILE_PATH,
+        constants.AB_EVALUATION_RESULT_FILE_PATH,
+    ]
+    df_list = [pd.read_csv(fp) for fp in file_paths]
+    combined_df = pd.concat(df_list, ignore_index=True)
+    combined_df = combined_df.iloc[:, 1:-1]
+    combined_df = combined_df[combined_df["Model"].str.contains("Tuned|^Naive$", regex=True)]
+    combined_df["MAPE (%)"] = combined_df["MAPE (%)"] * 100
+    combined_df["Model"] = combined_df["Model"].str.replace(" Tuned", "", regex=False)
+    combined_df["Model"] = combined_df["Model"].replace({"Naive": "Initial naïve"})
+    combined_df = combined_df.round(2)
+    combined_df.to_csv(constants.COMBINED_EVALUATION_RESULT_FILE_PATH)

### 2025-05-07 tool/constants.py

@@ -32,6 +32,8 @@ NETWORK_ROUTING_RESULT_FILE_PATH = (
 )
 
 TRAFFIC_CONTROL_FILE_PATH = OUTPUT_DIR + "table1_traffic_controls.csv"
+COMBINED_EVALUATION_RESULT_FILE_PATH = OUTPUT_DIR + "table2_combined_evaluation_result.csv"
+
 
 RF_PREDICTION_RESULT_FILE_PATH = OUTPUT_DIR + "rf_prediction_result.csv"
 GB_PREDICTION_RESULT_FILE_PATH = OUTPUT_DIR + "gb_prediction_result.csv"

### 2025-05-07 tool/constants.py
Refactored In instruction.py for more details
added step-by-step details for instruction.py

@@ -34,11 +34,10 @@ NETWORK_ROUTING_RESULT_FILE_PATH = (
 TRAFFIC_CONTROL_FILE_PATH = OUTPUT_DIR + "table1_traffic_controls.csv"
 COMBINED_EVALUATION_RESULT_FILE_PATH = OUTPUT_DIR + "table2_combined_evaluation_result.csv"
 
-
-RF_PREDICTION_RESULT_FILE_PATH = OUTPUT_DIR + "rf_prediction_result.csv"
-GB_PREDICTION_RESULT_FILE_PATH = OUTPUT_DIR + "gb_prediction_result.csv"
-AB_PREDICTION_RESULT_FILE_PATH = OUTPUT_DIR + "ab_prediction_result.csv"
-DT_PREDICTION_RESULT_FILE_PATH = OUTPUT_DIR + "dt_prediction_result.csv"
+BEST_RF_RANDOM_PARAM_FILE_PATH = OUTPUT_DIR + "best_rf_random_param.csv"
+BEST_GB_RANDOM_PARAM_FILE_PATH = OUTPUT_DIR + "best_gb_random_param.csv"
+BEST_AB_RANDOM_PARAM_FILE_PATH = OUTPUT_DIR + "best_ab_random_param.csv"
+BEST_DT_RANDOM_PARAM_FILE_PATH = OUTPUT_DIR + "best_dt_random_param.csv"
 
 RF_EVALUATION_RESULT_FILE_PATH = OUTPUT_DIR + "rf_evaluation_result.csv"
 GB_EVALUATION_RESULT_FILE_PATH = OUTPUT_DIR + "gb_evaluation_result.csv"
@@ -46,7 +45,3 @@ AB_EVALUATION_RESULT_FILE_PATH = OUTPUT_DIR + "ab_evaluation_result.csv"
 DT_EVALUATION_RESULT_FILE_PATH = OUTPUT_DIR + "dt_evaluation_result.csv"
 NETWORK_ROUTING_EVALUATION_RESULT_FILE_PATH = OUTPUT_DIR + "naive_evaluation_result.csv"
 
-BEST_RF_RANDOM_PARAM_FILE_PATH = OUTPUT_DIR + "best_rf_random_param.csv"
-BEST_GB_RANDOM_PARAM_FILE_PATH = OUTPUT_DIR + "best_gb_random_param.csv"
[…]

### 2025-05-07 tool/constants.py
pre-commit whitespace fix

@@ -44,4 +44,3 @@ GB_EVALUATION_RESULT_FILE_PATH = OUTPUT_DIR + "gb_evaluation_result.csv"
 AB_EVALUATION_RESULT_FILE_PATH = OUTPUT_DIR + "ab_evaluation_result.csv"
 DT_EVALUATION_RESULT_FILE_PATH = OUTPUT_DIR + "dt_evaluation_result.csv"
 NETWORK_ROUTING_EVALUATION_RESULT_FILE_PATH = OUTPUT_DIR + "naive_evaluation_result.csv"
-

### 2025-05-10 paper/main.tex
update paper

@@
 As we shall show in this article, across the Los Angeles urban area, our average trip's travel time prediction differs from Google's \enquote{gold standard} prediction by only 0.38 seconds.
-In contrast, simple---but common in the literature---naïve travel time prediction minimizing max-speed edge traversal times under-predicts by 183.69 seconds on average.
+In contrast, simple---but common in the literature---naïve travel time prediction minimizing max-speed edge traversal times under-predicts by 183.68 seconds on average.
 
@@
     \centering
-    \includegraphics[width=1.0\textwidth]{fig_workflow.jpg}
+    \includegraphics[width=1.0\textwidth]{fig_workflow_larger_font.jpg}
     \caption{Detailed model workflow.}\label{fig:workflow}
@@
         \midrule
-        Initial naïve       & 21.15 & 183.69 & 48214.06 &  -182.85 & <0.01 & 0.79 & 0.74  \\
-        \textbf{Random forest}       &  \textbf{8.42} &  \textbf{75.32} & \textbf{12154.79} &     \textbf{0.38} &  \textbf{0.76} & \textbf{1.01} & \textbf{0.93}  \\
+        Initial naïve       & 21.15 & 183.68 & 48214.06 &  -182.85 & <0.01 & 0.79 & 0.74  \\
+        \textbf{Random forest}       &  \textbf{8.41} &  \textbf{75.32} & \textbf{12154.79} &     \textbf{0.38} &  \textbf{0.76} & \textbf{1.01} & \textbf{0.93}  \\
         Gradient boosting   &  7.86 &  71.99 & 12630.03 &   -19.20 & <0.01 & 0.99 & 0.93  \\
@@
 
-The data that support the findings of this study are available at [OFFICIAL REPOSITORY URL REDACTED FOR ANONYMOUS PEER REVIEW.\@ REVIEWERS, PLEASE USE THIS ANONYMIZED URL FOR YOUR REVIEW INSTEAD:\@ https://figshare.com/s/0acd076b1a11286ff3e4].
+The code and data that support the findings of this study are available at [REDACTED].
+Google's terms of use forbid publicly redistributing their data: for reproducibility purposes, those training data are available directly from the authors upon request.
+[Reviewers: please use this anonymized URL for your review: https://figshare.com/s/0acd076b1a11286ff3e4 We have submitted the training data file separately to the journal as supplementary material for review only, to abide by the terms of use.]
 %https://github.com/gboeing/travel-time-prediction

### 2025-05-13 paper/main.tex
update statement

@@
     \centering
-    \includegraphics[width=1.0\textwidth]{fig_workflow_larger_font.jpg}
+    \includegraphics[width=1.0\textwidth]{fig_workflow.jpg}
     \caption{Detailed model workflow.}\label{fig:workflow}
@@
 
-The code and data that support the findings of this study are available at [REDACTED].
-Google's terms of use forbid publicly redistributing their data: for reproducibility purposes, those training data are available directly from the authors upon request.
-[Reviewers: please use this anonymized URL for your review: https://figshare.com/s/0acd076b1a11286ff3e4 We have submitted the training data file separately to the journal as supplementary material for review only, to abide by the terms of use.]
-%https://github.com/gboeing/travel-time-prediction
+The code and data that support the findings of this study are available at [REDACTED], minus one training data file which Google's terms of use forbid publicly redistributing.
+For replication purposes, data equivalent to those in that file are freely available (up to a monthly usage quota) directly from the Google Routes API (\url{https://developers.google.com/maps/documentation/routes}) by signing up for an API key from the Google Cloud Console (\url{https://console.cloud.google.com/}).
+%\url{https://github.com/gboeing/travel-time-prediction}
+
+[Reviewers: please use this anonymized URL for your review instead: \url{https://figshare.com/s/0acd076b1a11286ff3e4}.
+We have submitted the original training data file separately to the journal directly as supplementary material for review only, to abide by the terms of use.]
 
